---
title: "Exporting custom columns created in scenarios"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exporting custom columns created in scenarios}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## What this is for

A scenario function can attach new columns to the synthetic population
(`synthpop$pop`). By default those columns are **dropped** before the
lifecourse is written to disk, so they never reach the summaries or tables.

This vignette explains the mechanism that lets a user-created column survive to
the lifecourse output **and** be aggregated into the summary files and tables,
with no changes to the model's source code. The mechanism is purely a **naming
convention** plus, optionally, one entry in the design YAML.

## The two requirements

For a scenario-created column to appear in the summaries, it must satisfy
**both** of the following:

1. **Be kept by the lifecourse pruning step** — either by being listed in
   `cols_for_output` in the design YAML, *or* by being named with a recognised
   suffix that is kept automatically (`_prvl`, `_contd`, `_costs`, `_dgns`,
   `_mrtl`, or a `cms_` prefix).
2. **Use a recognised suffix** so the export step knows how to aggregate it.

The recognised suffixes and their treatment are:

| Suffix    | Interpreted as        | Aggregation                              | Output folder(s)                                   |
|-----------|-----------------------|------------------------------------------|----------------------------------------------------|
| `_prvl`   | duration counter      | prevalence `SUM(wt)` where `col > 0` **and** incidence `SUM(wt)` where `col = 1` | `summaries/prvl_scaled_up/`, `summaries/incd_scaled_up/` (+ `_esp`) |
| `_contd`  | continuous quantity   | population-weighted mean                  | `summaries/contd_scaled_up/`, `summaries/contd_esp/` |
| `_costs`  | monetary cost         | `SUM(col * wt)`                           | `summaries/costs_scaled_up/`, `summaries/costs_esp/` |

`_prvl` and `_costs` ride on the existing exporters; `_contd` is exported by a
dedicated step that you opt into with `type = c(..., "contd")` in
`export_summaries()`.

> **Note on the `_costs` suffix.** The model's *built-in* cost columns use the
> singular `_cost` suffix (`healthcare_cost`, `socialcare_cost`,
> `informalcare_cost`, `total_cost`, …). Custom cost columns use the **plural**
> `_costs` suffix so the two never collide. Both end up side-by-side in the
> `costs` summary.

## How `_prvl` yields both prevalence and incidence

The disease prevalence columns in the model are not 0/1 flags — they are
**duration counters**: the value is `1` in the year the condition first appears
(the incidence event), then `2`, `3`, … for each subsequent year the person
carries the condition. The prevalence exporter counts everyone with `col > 0`;
the incidence exporter counts only `col = 1`. A single column drives both.

To build such a counter from a 0/1 event vector, use `carry_forward_incr()`, the
same helper the model uses internally:

```{r}
# inside a scenario function, `synthpop$pop` is keyed by (pid, year)
synthpop$pop[, my_event_prvl := carry_forward_incr(
  as.integer(some_condition),  # 1 in the event year, 0 otherwise
  pid_mrk,                     # per-simulant marker (already present)
  recur = FALSE,               # FALSE = once-only (e.g. CHD); TRUE = can recur
  y = 1L,
  byref = TRUE
)]
```

If you only care about prevalence and not incidence timing, a plain
non-decreasing indicator also works — anything `> 0` counts as prevalent and
anything `= 1` counts as incident in that year.

## A worked example

The example below defines a primary-prevention scenario that, from 2025 onwards,
reduces systolic blood pressure (SBP) by 10% and records three custom columns:

* `sbp_intervention_prvl` — a flag for "is currently exposed to the
  intervention", as a duration counter so it yields both a prevalence and an
  incidence summary;
* `sbp_excess_contd` — a continuous quantity (mmHg of SBP above 130), summarised
  as a population-weighted mean;
* `sbp_intervention_costs` — a per-person-year monetary cost of delivering the
  intervention.

> **Year format.** In England the lifecourse `year` column is stored in *short*
> form (e.g. `25` for 2025), so scenario conditions use `year >= 25L`. The
> tables produced by `$export_tables()` are converted to full years (2025).

### Step 1 — (optionally) declare the columns in the design YAML

`_prvl`, `_contd` and `_costs` columns are kept automatically by the pruning
step, so strictly you do not need to list them. Listing them anyway is good
documentation and is **required** for any custom column that does *not* use a
recognised suffix:

```yaml
cols_for_output:
- pid
- pid_mrk
- year
- mc
- age
- agegrp
- sex
- dimd
- wt
- wt_esp
- sbp_intervention_prvl    # optional for suffixed cols, but explicit is nice
- sbp_excess_contd
- sbp_intervention_costs
```

### Step 2 — create the columns in the scenario function

Create the columns in **every** scenario (including the baseline) so the
lifecourse schema is consistent across scenarios — set them to their "off"
values in the baseline.

```{r}
source("./global.R")
IMPACTncd <- Simulation$new("./inputs/sim_design.yaml")

# Baseline scenario (sc0): intervention OFF, but columns present at base values.
IMPACTncd$update_primary_prevention_scn(
  function(synthpop) {
    synthpop$pop[, sbp_intervention_prvl := 0L]
    synthpop$pop[, sbp_excess_contd := pmax(0, sbp_curr_xps - 130)]
    synthpop$pop[, sbp_intervention_costs := 0]
    NULL
  }
)

IMPACTncd$
  del_logs()$
  del_outputs()$
  run(1:2, multicore = TRUE, "sc0")   # baseline

# Intervention scenario (sc1): 10% SBP reduction from 2025 onwards.
IMPACTncd$update_primary_prevention_scn(
  function(synthpop) {
    synthpop$pop[year >= 25L, sbp_curr_xps := sbp_curr_xps * 0.9]

    # (1) _prvl: duration counter -> drives prvl AND incd summaries
    synthpop$pop[, sbp_intervention_prvl := carry_forward_incr(
      as.integer(year >= 25L),
      pid_mrk,
      recur = FALSE,
      y = 1L,
      byref = TRUE
    )]

    # (2) _contd: continuous -> population-weighted mean
    synthpop$pop[, sbp_excess_contd := pmax(0, sbp_curr_xps - 130)]

    # (3) _costs: monetary -> SUM(col * wt); feeds net_costs downstream
    synthpop$pop[, sbp_intervention_costs := fifelse(year >= 25L, 500, 0)]

    NULL
  }
)

IMPACTncd$
  run(1:2, multicore = TRUE, "sc1")   # intervention scenario
```

### Step 3 — export summaries, including the continuous one

`prvl`, `incd` and `costs` are standard types and pick up the custom columns
with no extra work. Add `"contd"` to also export the weighted-mean summary for
`_contd` columns.

```{r}
IMPACTncd$export_summaries(
  multicore = TRUE,
  type = c(
    "le", "hle", "dis_char",
    "prvl", "incd", "dis_mrtl", "mrtl",
    "all_cause_mrtl_by_dis", "cms",
    "qalys", "costs",
    "contd"            # <- exports summaries/contd_scaled_up/ etc.
  )
)
```

## What the outputs look like

Each summary is one Parquet dataset per Monte Carlo iteration, stratified by
`strata_for_output` (by default `scenario, year, agegrp, sex, dimd`) plus `mc`
and a `popsize` column. The idiomatic way to read a summary is
`arrow::open_dataset()` on the summary *folder*, which combines all Monte Carlo
files at once.

```{r}
library(arrow)
library(data.table)

smr <- file.path(IMPACTncd$design$sim_prm$output_dir, "summaries")

# (1) prevalence of the intervention flag
prvl <- as.data.table(open_dataset(file.path(smr, "prvl_scaled_up")))
prvl[, .(mc, scenario, year, agegrp, sex, dimd, popsize, sbp_intervention_prvl)]

# the SAME _prvl column also produced an incidence column (col == 1)
incd <- as.data.table(open_dataset(file.path(smr, "incd_scaled_up")))
incd[, .(mc, scenario, year, agegrp, sex, dimd, sbp_intervention_incd)]

# (2) weighted-mean of the continuous column
contd <- as.data.table(open_dataset(file.path(smr, "contd_scaled_up")))
contd[, .(mc, scenario, year, agegrp, sex, dimd, popsize, sbp_excess_contd)]

# (3) summed cost column, alongside the model's built-in cost columns
costs <- as.data.table(open_dataset(file.path(smr, "costs_scaled_up")))
costs[, .(mc, scenario, year, agegrp, sex, dimd, sbp_intervention_costs)]
```

Notes on the column names:

* The `_prvl` column keeps its name in the prevalence file
  (`sbp_intervention_prvl`) and is renamed `…_incd` in the incidence file
  (`sbp_intervention_incd`).
* The `_contd` column keeps its name and holds the weighted mean.
* The `_costs` column keeps its name and holds the weighted sum.

## Downstream tables (`$export_tables()`)

`$export_tables()` turns the per-Monte-Carlo summaries into quantile tables in
`output_dir/tables/`. It selects columns by **suffix**, so most custom columns
are tabulated automatically — you do not need to edit any code:

* `_prvl` columns feed the `prevalence`, `prevalence change` and
  `case-years prevented or postponed` (cypp) tables.
* the derived `_incd` columns feed `incidence`, `incidence change` and
  `cases prevented or postponed` (cpp).
* `_costs` columns feed `costs` and `net costs`.
* `_contd` columns feed the `continuous outcome` tables (a population-weighted
  mean, re-weighted by `popsize` when collapsing strata).

```{r}
IMPACTncd$export_tables()
```

### Costs, net costs and economic evaluation

The `net costs` output is the difference versus the comparator scenario (by
default `sc0`). Because it discovers all cost columns by suffix, your custom
cost appears in net-cost tables without further configuration.

Incremental cost-effectiveness ratios (ICER) and net monetary benefit (NMB) are
produced directly by `$export_tables()` when `cea = TRUE` (the default). It
writes `cost-effectiveness by <strata> (<perspective>-<scale>) (not
standardised).csv` files containing the cumulative discounted incremental QALYs
(`dQALYs_cuml`) and costs (`dCosts_cuml`), the `ICER`, and the `NMB` at each
willingness-to-pay threshold, for the EQ-5D-5L QALY scale and from two
perspectives:

* **societal** — uses `total_cost` (healthcare + social care + informal care,
  net of `economic_output`), plus all custom `_costs` columns;
* **healthcare** — uses `healthcare_cost` (direct treatment costs only).

Discounting is applied with separate annual rates for costs and QALYs
(`cost_discount_rate`, `qaly_discount_rate`, default 3.5% each per NICE) and is
controlled **only** by these `export_tables()` arguments — there is no
`discounting` block in `sim_design.yaml`. Discounting is applied **only** to the
cost-effectiveness tables; the prevalence, incidence, mortality, QALY and cost
tables are reported undiscounted. `discount_from_year` defaults to the baseline
(reference) year. The willingness-to-pay thresholds are set with the `wtp`
argument (default `c(20000, 30000)` GBP/QALY, the NICE thresholds):

```{r}
IMPACTncd$export_tables(
  cea = TRUE,
  wtp = c(20000, 30000),       # GBP per QALY
  qaly_discount_rate = 3.5,    # discounting lives only in the CEA tables
  cost_discount_rate = 3.5
)
```

**Where do your custom `_costs` columns count?** Custom cost columns are
**always** added to the **societal** perspective. By default they are *not*
included in the healthcare perspective (which captures direct treatment costs
only). If some of your custom costs *are* healthcare costs, pass their column
names to `custom_costs_in_healthcare` in `export_tables()` and they are added to
the healthcare perspective as well, e.g.
`custom_costs_in_healthcare = c("screening_costs", "drug_costs")`. For example,
an intervention-delivery cost paid by the health system should be listed here,
whereas a productivity-loss or out-of-pocket cost should be left in the societal
perspective only. (Passing `TRUE` includes *all* custom cost columns and
`NULL`/`FALSE` includes none.)

If you need a bespoke calculation, you can still derive ICER and NMB yourself
from the `net costs` and `net QALYs` tables in your own analysis scripts.

## Tips and gotchas

* **Keep the column present in every scenario.** If a column exists in `sc1` but
  not in `sc0`, the baseline rows will be missing it. When the lifecourse files
  are combined across scenarios the missing values become `NULL`, which the
  prevalence/incidence/cost aggregations treat as 0. If you want a genuine zero
  baseline, create the column in *every* scenario (including the baseline) — e.g.
  set it to `0` when the intervention is not active.
* **Avoid the substring `wt` in custom column names.** The ESP-weighted variants
  of some summaries are produced by string-replacing `wt` with `wt_esp` in the
  SQL. A column name containing `wt` could be corrupted by that substitution.
* **`_contd` is opt-in.** Remember to add `"contd"` to the `type` argument of
  `export_summaries()`, otherwise the continuous summary is not written.
* **Type matters.** `_prvl` and `_incd` logic compares against integers (`> 0`,
  `= 1`), so make `_prvl` columns integer counters. `_contd` and `_costs` should
  be numeric.
