---
title: "Using your own exposure distribution tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using your own exposure distribution tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## What this is for

Every risk factor the model generates for the synthetic population is driven by
a **distribution table** — a partitioned parquet dataset holding the fitted
distribution parameters (e.g. `mu`, `sigma`, `nu`, `tau` for a continuous
exposure) by strata (age, sex, deprivation, region, ...). These tables ship with
the model under `./inputs/exposure_distributions/` and are referenced from the
`exposure_definitions` block of your `sim_design*.yaml`:

```yaml
exposure_definitions:
  - name: bmi
    file_name: bmi        # -> ./inputs/exposure_distributions/bmi
    var_name: bmi
    rank_var: rank_bmi
    distribution: continuous
    qfun: fqBCPEo
    join_fn: lookup_dt
    min_value: 14
    max_value: 70
```

This vignette shows how to point any exposure at **your own** table — for
example a locality-specific refit, a sensitivity analysis, or a "what if the
BMI distribution looked like *this*" experiment — **without editing the model's
source or its bundled inputs**, and in a way that works both on your own
machine and inside the Docker image.

> **Scope.** This is about *re-pointing* the table behind an exposure the model
> already knows about. Adding a brand-new exposure *variable* is a larger task
> (it also touches the correlation structure, the relative-risk files and the
> C++ engine) and is out of scope here.

## How `file_name` is resolved

`file_name` is resolved with three rules (implemented in
`Design$resolve_exposure_path()`):

| `file_name` looks like            | Resolved to                                              | Use it for                          |
|-----------------------------------|----------------------------------------------------------|-------------------------------------|
| a **bare name**, e.g. `bmi`       | `./inputs/exposure_distributions/bmi` (the stock table)  | the default; unchanged behaviour    |
| a **relative path with `/`**, e.g. `tables/bmi_local` or `./bmi_local` | resolved **against the directory of your `sim_design*.yaml`** | shipping a custom table with a scenario |
| an **absolute path**, e.g. `/mnt/data/bmi` | used verbatim                                     | large tables on external storage    |

The middle rule is the recommended one: put your table next to your scenario
files and reference it with a relative path. Because the path is resolved
against the **YAML's own directory**, the same config works unchanged on the
host and inside the container (the setup scripts bind-mount your scenario folder
at `/IMPACTncd_England/scenarios`, so no path translation is needed).

## Worked example: a custom BMI table

### 1. Lay out a scenario folder

```
my_scenario/
├── sim_design_my.yaml
├── simulate_my.R
└── tables/
    └── bmi_local/            # your partitioned parquet dataset
        ├── year=13/part-0.parquet
        ├── year=14/part-0.parquet
        └── ...
```

### 2. Create the custom table with the right schema

Your table must have the **same columns** as the stock table it replaces: the
join keys (the strata the model matches on) plus the distribution
parameters/thresholds. The simplest way to get the schema right is to start from
the stock table and modify it. For example, to shift the whole BMI distribution
upwards by 10%:

```{r}
library(IMPACTncdEngland) # provides read_parquet_dt + fpBCPEo via CKutils
library(data.table)

bmi <- read_parquet_dt("./inputs/exposure_distributions/bmi")
names(bmi)
#> "age" "ethnicity" "qimd" "sex" "sha" "smok_status"
#> "mu" "sigma" "nu" "tau" "maxq" "minq" "year"

bmi[, mu := mu * 1.10]                       # shift the distribution

# IMPORTANT: recompute minq/maxq for the modified parameters (see note below).
# fpBCPEo is the model's probability function for this exposure (from qfun
# fqBCPEo); min_value/max_value (14/70) come from the exposure definition.
bmi[, minq := fpBCPEo(rep(14, .N), mu, sigma, nu, tau)]
bmi[, maxq := fpBCPEo(rep(70, .N), mu, sigma, nu, tau)]

arrow::write_dataset(
  bmi,
  "my_scenario/tables/bmi_local",
  format = "parquet",
  partitioning = "year"
)
```

> **`minq`/`maxq` are required for continuous exposures.** They are the cached
> quantile bounds that map a uniform rank onto the truncated
> `[min_value, max_value]` range. The synthpop generator will recompute them in
> memory if they are missing, **but** the PARF / initial-prevalence code reads
> them straight from the table — so a custom continuous table that omits them
> (or ships bounds that don't match its parameters) will error or produce
> inconsistent starting values. Whenever you change a distribution parameter
> (`mu`, `sigma`, `nu`, `tau`), recompute `minq`/`maxq` as shown above, using the
> exposure's own probability function (`fpBCPEo` here; in general the `q` in the
> exposure's `qfun` becomes `p`, e.g. `fqBCT` → `fpBCT`). If you leave the
> distribution parameters untouched, keep the stock `minq`/`maxq` as they are.

### 3. Point the exposure at your table

In `sim_design_my.yaml`, change only the `file_name`:

```yaml
  - name: bmi
    file_name: tables/bmi_local   # <- YAML-relative path to your table
    var_name: bmi
    rank_var: rank_bmi
    distribution: continuous
    qfun: fqBCPEo
    join_fn: lookup_dt
    min_value: 14
    max_value: 70
```

### 4. Run as usual

```{r}
source("./global.R")
IMPACTncd <- Simulation$new("my_scenario/sim_design_my.yaml")

# Optional: confirm the exposure resolved to your table
IMPACTncd$design$exposures$bmi$file_path
#> ".../my_scenario/tables/bmi_local"

IMPACTncd$
  del_logs()$
  del_outputs()$
  run(1:2, multicore = TRUE, "sc0")$
  export_summaries(multicore = TRUE)
```

Nothing else changes. Every other exposure keeps reading its stock table.

## Using it from Docker

The user setup script bind-mounts your scenario folder into the container:

```bash
./docker_setup/setup_user_docker_env.sh -Tag main \
  -ScenariosDir /path/to/my_scenario \
  -SimDesignYaml /path/to/my_scenario/sim_design_my.yaml
```

Because `tables/bmi_local` is resolved relative to the mounted YAML, it points
at your table inside the container with no further configuration — the same YAML
that works on your laptop works in the container.

## It stays consistent everywhere

A re-pointed table is used **everywhere** the model reads that exposure — both
when generating the synthetic population and when generating the PARF
(Population Attributable Risk Fraction) starting points. Internally the model
routes those reads through `Design$xps_table_path(<exposure name>)`, which
returns your table's location. So there is no risk of the synthpop using your
table while some other stage silently falls back to the stock one.

## Getting the schema right

- **Continuous** exposures (`distribution: continuous`) need the join-key
  columns plus the parameters your `qfun` expects (commonly `mu`, `sigma`, `nu`,
  `tau`) **and** the `minq`/`maxq` bounds (see the note in step 2 — required by
  the PARF/initial-prevalence path, and must match your parameters).
- **Ordinal** exposures need the join-key columns plus the threshold columns.
- **Binary** exposures need the join-key columns plus the probability column.

Open the stock table to see exactly what is expected:

```{r}
str(read_parquet_dt("./inputs/exposure_distributions/bmi"))
# or, columns only, without reading all the data:
names(arrow::open_dataset("./inputs/exposure_distributions/bmi"))
```

Keep the join-key columns and their types identical to the stock table;
mismatched keys will silently fail to join and produce missing values.

## Quick reference

- Bare name  → stock table (default, backward compatible).
- `path/with/slash` → relative to the `sim_design*.yaml` directory (recommended
  for custom tables; Docker-friendly).
- `/absolute/path` → used as-is.
- The custom table must match the stock table's schema for that exposure.
- Re-pointing is honoured in both synthpop and PARF generation.

See also `vignette("how_to_run_scenarios")` for the broader scenario workflow.
