---
title: "Understanding the Inputs Manifest System"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding the Inputs Manifest System}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

The IMPACTncd simulation relies on thousands of input files ظ¤ relative risk
tables, population projections, exposure distributions, mortality data, and
disease burden statistics. When any of these files changes, the simulation
must regenerate the downstream artifacts that depend on them: synthetic
populations (synthpops), compiled relative risk (RR) `.fst` files, and
Population Attributable Risk Fraction (PARF) files.

The **Inputs Manifest System** provides a unified mechanism to:

1. **Track** all input files with their checksums
2. **Detect** which files have changed since the last simulation run
3. **Map** file changes to affected downstream artifacts using the model's
   causal structure
4. **Trigger** automatic regeneration of stale artifacts

This vignette explains the architecture, data flow, and design decisions to
help new developers understand and extend the system.

## Architecture

### The problem: fragmented tracking

Before the manifest system, the project used three separate, inconsistent
mechanisms:

| Mechanism | What it tracked | Weaknesses |
|-----------|----------------|------------|
| `simulation/fileversion.csv` | Compiled RR `.fst` files and PARF snapshots | Absolute paths (non-portable); shared file with race conditions; no synthpop tracking |
| Synthpop meta YAML files | Design parameters only | Did not track input data files ظ¤ synthpops were never regenerated when upstream data changed |
| `simulation/zenodo_manifest.csv` | Zenodo upload state | Completely separate system |

### The solution: unified manifest

The `InputsManifest` R6 class replaces all three with a single CSV manifest
(`simulation/inputs_manifest.csv`) that records every data file in `inputs/`
with its:

- **Relative path** (portable across machines)
- **xxhash64 checksum** (fast and consistent)
- **File size** and **modification time** (for quick change detection)
- **Category** (top-level subdirectory: `RR`, `exposure_distributions`, etc.)
- **Zenodo-managed flag** (whether the file is uploaded to Zenodo or tracked
  by git)

## Data flow

The following diagram shows how the manifest integrates with the simulation
pipeline:

```
 inputs/                   simulation/
 ظ¤£ظ¤ظ¤ RR/                   ظ¤£ظ¤ظ¤ inputs_manifest.csv      ظ manifest
 ظ¤£ظ¤ظ¤ exposure_dist/        ظ¤£ظ¤ظ¤ rr_generation_record.csv  ظ tracks RR compilations
 ظ¤£ظ¤ظ¤ mortality/            ظ¤£ظ¤ظ¤ parf_generation_record.csv ظ tracks PARF generations
 ظ¤£ظ¤ظ¤ pop_projections/      ظ¤£ظ¤ظ¤ rr/                       ظ compiled RR .fst files
 ظ¤£ظ¤ظ¤ pop_estimates_lsoa/   ظ¤£ظ¤ظ¤ parf/                     ظ PARF .qs files
 ظ¤¤ظ¤ظ¤ disease_burden/       ظ¤¤ظ¤ظ¤ synthpop/                  ظ synthpop .fst files

 Simulation$new()
   ظ¤é
   ظ¤£ظ¤ InputsManifest$new() ظ¤ظ¤ظû║ load or generate manifest
   ظ¤é
   ظ¤£ظ¤ detect_changes(design) ظ¤ظ¤ظû║ compare filesystem vs manifest
   ظ¤é    ظ¤é
   ظ¤é    ظ¤£ظ¤ affected_synthpops?  ظ¤ظ¤ظû║ synthpop checksums change ظ¤ظ¤ظû║ auto-regenerate
   ظ¤é    ظ¤£ظ¤ affected_rr?         ظ¤ظ¤ظû║ delete stale .fst files  ظ¤ظ¤ظû║ auto-regenerate
   ظ¤é    ظ¤¤ظ¤ affected_diseases?   ظ¤ظ¤ظû║ delete stale PARFs       ظ¤ظ¤ظû║ auto-regenerate
   ظ¤é
   ظ¤¤ظ¤ generate() ظ¤ظ¤ظû║ re-scan and save updated manifest
```

### Step-by-step: what happens on `Simulation$new()`

1. An `InputsManifest` object is created, pointing at `./inputs/` and
   `./simulation/inputs_manifest.csv`.

2. **First run** (no manifest exists): `generate()` scans all input files,
   computes their hashes, and saves the manifest. No artifacts are deleted.

3. **Subsequent runs** (manifest exists):
   a. The existing manifest is loaded
   b. `detect_changes(design)` compares the filesystem against the manifest
   c. If changes are found:
      - Stale compiled RR `.fst` files are deleted
      - Stale PARF files are deleted
      - The manifest is regenerated to reflect the current state
   d. Synthpop staleness is handled differently (see below)

4. The existing regeneration logic in `SynthPop`, `ExposureEffect`, and
   `Disease` classes detects that the files are missing and recreates them.

## Hashing strategy

### Why xxhash64?

The `inputs/exposure_distributions/` directory alone can contain 10,000+
parquet files. Hashing all of them with MD5 would take minutes. `xxhash64`
(via `digest::digest(file = path, algo = "xxhash64")`) is ~10x faster
while still providing reliable change detection for our use case (detecting
accidental or intentional data file changes, not cryptographic security).

### The mtime pre-filter

Re-hashing 10,000 unchanged files on every simulation start would be
wasteful. The `compare()` method uses **modification time (mtime)
pre-filtering**:

1. For each file in the manifest, check if the filesystem mtime matches the
   stored mtime
2. If mtime is unchanged ظْ assume the file is unchanged (fast path)
3. If mtime differs ظْ re-hash the file and compare against the stored hash
4. This reduces incremental checks from O(n ├ù hash_time) to
   O(n ├ù stat_time + m ├ù hash_time), where `m` is the number of actually
   modified files

### Parallel hashing

When `parallel = TRUE` is passed to `generate()`, hashing is distributed
across PSOCK cluster workers using `parallelly::makeClusterPSOCK()` +
`parallel::parLapplyLB()`. The standalone `.manifest_hash_worker()` function
(defined outside the R6 class) handles the hashing on each worker.

**Why PSOCK and not forking?** The project avoids `mclapply`/`doParallel`
because `zip::zip()` uses C internals that are not fork-safe, and R6 objects
with HTTP connections (e.g., `zen4R::ZenodoManager`) cannot be serialized to
forked children. PSOCK clusters serialise the function and its data to
independent R processes, avoiding these issues.

## Causal invalidation

The most sophisticated part of the system is mapping file changes to affected
downstream artifacts. This is done by `detect_changes()`.

### Category mapping

Each file's **category** is its top-level subdirectory in `inputs/`:

| Category | Affects |
|----------|---------|
| `pop_estimates_lsoa` | Synthpops |
| `exposure_distributions` | Synthpops |
| `mortality` | Synthpops |
| `pop_projections` | Synthpops |
| `RR` | Compiled RR `.fst` files and downstream PARFs |
| `disease_burden` | PARF files (data files only, not `.yaml`) |

### Causal graph traversal

When an RR `.csvy` file changes (e.g., `RR/bmi~chd.csvy`), the system:

1. Parses the `exposure~outcome` from the filename ظْ `bmi~chd`
2. Identifies `chd` as the directly affected disease
3. Checks `design$RR` names for any edges where `chd` is the *exposure*
   (e.g., `chd~hf`)
4. Uses **breadth-first search (BFS)** to find all transitively affected
   diseases (e.g., `hf`)
5. Deletes PARF files for all affected diseases

This ensures that when a root exposure changes, all downstream diseases in
the causal chain are correctly regenerated.

### What does NOT trigger invalidation

- **`disease_burden/*.yaml` files**: These are informational metadata and are
  tracked in the manifest but do not trigger PARF regeneration. Only actual
  data files (`.csv`, `.fst`, etc.) in `disease_burden/` trigger
  invalidation.
- **`sim_design*.yaml` files**: Design parameter changes are already handled
  by the existing design-parameter checksums in the SynthPop class.

## How synthpop staleness works

Synthpop files embed their checksum in the filename:
`synthpop_<checksum>_<mc>.fst`. The checksum is computed by
`SynthPop$gen_checksum()` from:

- Design parameters (simulation years, age ranges, etc.)
- The list of LSOAs
- The **synthpop input hash** from `InputsManifest$get_synthpop_input_hash()`

The synthpop input hash is an aggregate of all file hashes in the
`pop_estimates_lsoa`, `exposure_distributions`, `mortality`, and
`pop_projections` categories. When any file in these categories changes:

1. The aggregate hash changes
2. The synthpop checksum changes
3. The synthpop filename changes
4. The old synthpop file is simply not found by the simulation
5. `gen_synthpop()` runs automatically (existing logic)

**Backward compatibility**: When no manifest exists (very first run before
`Simulation$new()` creates one), the input hash is `""` and the checksum
behaves exactly as before the manifest system was introduced.

## How RR and PARF staleness works

### Compiled RR `.fst` files

Each `ExposureEffect` object compiles a `.csvy` source file into optimised
`.fst` files (`rr_<exposure>~<outcome>_l.fst` and `_indx.fst`). The
generation record (`simulation/rr_generation_record.csv`) tracks which
source file and source hash were used for each compilation.

When `check_registry_valid()` runs:
1. It reads the generation record
2. Computes the current hash of the source `.csvy` file
3. Compares against the stored `source_hash`
4. If they differ, the `.fst` files are regenerated

### PARF files

Similarly, `Disease` objects track their PARF generation in
`simulation/parf_generation_record.csv`. The PARF checksum incorporates:
- Design parameters
- RR file hashes (so RR changes cascade to PARFs)
- Lag structures and distribution parameters

When `Simulation$new()` detects that an RR file has changed, it proactively
deletes the affected PARF files. The `Disease` class then detects the missing
PARFs and regenerates them.

## Zenodo integration

Not all files in `inputs/` are uploaded to Zenodo. Some small auxiliary files
are tracked by git instead. The manifest distinguishes between the two via the
`zenodo_managed` column.

### Marking Zenodo directories

```{r zenodo-dirs}
manifest <- InputsManifest$new()
manifest$generate()
manifest$set_zenodo_dirs(c(
  "RR", "exposure_distributions", "disease_burden",
  "mortality", "pop_projections"
))
manifest$save()
```

### Comparing with last upload

After setting up Zenodo directories, you can check what changed since the
last upload:

```{r zenodo-compare}
# Assuming IMPACTncd is a Simulation object with Zenodo configured
manifest <- IMPACTncd$get_inputs_manifest()
diff <- IMPACTncd$zenodo_mgr$compare_with_inputs_manifest(manifest)

if (!is.null(diff)) {
  cat("Files changed since last upload:", length(diff$modified), "\n")
  cat("New files:", length(diff$added), "\n")
}
```

### Snapshotting on upload

After a successful Zenodo upload, call `snapshot_manifest_for_upload()` to
record the current state. This creates
`simulation/inputs_manifest_at_last_upload.csv`, which is used by subsequent
`compare_with_inputs_manifest()` calls.

## Key classes and files

| File | Class/Function | Role |
|------|---------------|------|
| `R/InputsManifest_class.R` | `InputsManifest` (R6) | Core manifest: scan, hash, compare, detect changes |
| `R/InputsManifest_class.R` | `.manifest_hash_worker()` | Standalone worker for PSOCK parallel hashing |
| `R/Simulation_class.R` | `Simulation$initialize()` | Creates manifest, detects changes, deletes stale artifacts |
| `R/Simulation_class.R` | `Simulation$get_inputs_manifest()` | Public accessor for the manifest object |
| `R/SynthPop_class.R` | `SynthPop$gen_checksum()` | Includes `get_synthpop_input_hash()` in synthpop checksum |
| `R/ExposureEffect_class.R` | Private registry methods | Track RR compilation via `rr_generation_record.csv` |
| `R/Disease_class.R` | Private registry methods | Track PARF generation via `parf_generation_record.csv` |
| `R/ZenodoManager_class.R` | `compare_with_inputs_manifest()` | Compare current inputs vs last Zenodo upload |
| `R/ZenodoManager_class.R` | `snapshot_manifest_for_upload()` | Save manifest snapshot at upload time |
| `R/Design_class.R` | `Design$initialize()` | Cleans orphaned RR files and legacy `fileversion.csv` |

## Tracking files generated by the manifest system

The following files are generated by the manifest system and are excluded
from git via `.gitignore`:

| File | Purpose |
|------|---------|
| `simulation/inputs_manifest.csv` | Current manifest of all input files |
| `simulation/inputs_manifest_at_last_upload.csv` | Snapshot from last Zenodo upload |
| `simulation/rr_generation_record.csv` | Tracks which source produced each compiled RR |
| `simulation/parf_generation_record.csv` | Tracks which sources produced each PARF |
| `simulation/fileversion.csv` | Legacy file (auto-deleted by Design on startup) |

## Design decisions for developers

### Why not re-hash output artifacts at startup?

Synthpop and PARF files can be several gigabytes. Re-hashing them at every
`Simulation$new()` would add minutes of startup time. Instead, checksums are
embedded in filenames. When upstream inputs change, the checksum changes, the
expected filename changes, and the old file is simply not found ظ¤ triggering
automatic regeneration without ever reading the old file.

The `verify_artifact()` method provides an opt-in diagnostic for when you
need to actually verify file integrity (e.g., before a Zenodo upload or after
copying files between machines).

### Why standalone worker functions?

R6 objects with environments, closures, and connections cannot be reliably
serialised to PSOCK cluster workers. The project convention is to define
standalone worker functions *outside* the R6 class that accept only plain
serialisable data (character strings, numbers, lists) and return plain lists.
See `.manifest_hash_worker()` and `.zenodo_archive_worker()` for examples.

### Why separate generation records?

The old `fileversion.csv` was a single shared file for both RR compilations
and PARF generations, leading to potential race conditions when multiple
processes wrote to it simultaneously. The new system uses two separate files:
`rr_generation_record.csv` and `parf_generation_record.csv`.

### Why not invalidate on `disease_burden/*.yaml` changes?

The `.yaml` files in `disease_burden/` are informational metadata (e.g.,
descriptions, citations) that do not affect PARF calculations. Only the
actual data files (incidence CSVs, fatality FSTs) are used in generation, so
only those trigger invalidation. This avoids unnecessary regeneration when
someone updates a description.

## Extending the system

### Adding a new input category

If you add a new subdirectory to `inputs/` (e.g., `inputs/cost_data/`):

1. Files are automatically tracked by the manifest (no code changes needed
   for basic tracking)
2. To make changes trigger artifact regeneration, update
   `InputsManifest$detect_changes()` to map the new category to the
   appropriate downstream artifacts
3. If the new category affects synthpops, add it to the
   `synthpop_categories` vector in `detect_changes()` and to the
   `categories` vector in `get_synthpop_input_hash()`

### Adding a new artifact type

If you add a new type of generated artifact:

1. Create a generation record CSV (like `rr_generation_record.csv`)
2. Add staleness checking and registry update methods to the relevant class
3. In `Simulation$initialize()`, add deletion logic for stale artifacts when
   changes are detected
4. Add the generation record file to `.gitignore`
