---
title: "Managing Input Data with Zenodo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Managing Input Data with Zenodo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

IMPACTncdEngland includes a robust pipeline for managing input data files
via [Zenodo](https://zenodo.org/), a general-purpose open repository hosted
by CERN. This enables:

- **Reproducibility**: DOI-based versioning ensures you always know which data
  version was used
- **Collaboration**: Easy sharing of input datasets across team members
- **Archiving**: Long-term preservation of input data with persistent
  identifiers
- **Version Control**: Track changes to input data over time

The pipeline uses the `zen4R` package to interact with the Zenodo REST API,
and supports **parquet**, **fst**, and **qs** file formats commonly used in
the simulation. It can archive both **input data** (exposure distributions,
disease burden, mortality, etc.) and **simulation outputs** (PARFs, compiled
RR tables).

### What is a DOI?

A **Digital Object Identifier (DOI)** is a permanent, unique identifier for a
digital resource. When you publish data on Zenodo, it receives a DOI that will
always resolve to that exact version of the data, even decades later. This is
the standard way to cite research data in academic publications.

Zenodo provides two types of DOIs:

| Type | Example | Resolves to |
|------|---------|-------------|
| **Concept DOI** | `10.5281/zenodo.12345` | Always the *latest* published version |
| **Version DOI** | `10.5281/zenodo.12346` | A *specific* version (immutable) |

Use the **concept DOI** in your code and share it with collaborators. Cite the
**version DOI** in publications for exact reproducibility.

## Prerequisites

### 1. Install Required Packages

The following packages are required (they are Suggests in the package):

```{r install-packages}
# zen4R is the only required Zenodo dependency declared in Suggests.
# httr2 is used internally by the API fallback path (e.g. listing draft
# files) but is not declared in DESCRIPTION — install it explicitly.
install.packages(c("zen4R", "httr2"))
# Optional: for secure token storage
install.packages("keyring")
```

(`zip` and `digest` are already pulled in as package Imports — no need to
install them separately.)

### 2. Get a Zenodo Access Token (uploaders only)

> **Downloading the published data needs no account or token.** The IMPACTncd
> England input-data record is public, so `zenodo_connect()` followed by
> `zenodo_download_all()` works anonymously. You only need a token to
> **upload** or **publish** data. If you only want to run the model, skip to
> [Quick Start](#quick-start).

1. Create a Zenodo account at [zenodo.org](https://zenodo.org/)
2. Go to
   [Applications > Personal access tokens](https://zenodo.org/account/settings/applications/tokens/new/)
3. Create a new token with **deposit:write** and **deposit:actions** scopes
4. Save the token securely (see below)

**For testing**, create a separate sandbox account and token at
[sandbox.zenodo.org](https://sandbox.zenodo.org/account/settings/applications/tokens/new/).
Sandbox and production are **completely separate systems** — they do not
share accounts, tokens, or data. See
[Sandbox vs Production](#sandbox-vs-production) for details.

### 3. Secure Token Management

Your Zenodo token grants full access to your Zenodo deposits. **Never commit
tokens to version control!**

#### Option A: `pass` — GPG-Encrypted (Recommended for Linux/macOS)

[`pass`](https://www.passwordstore.org/) is a standard Unix password manager
that stores each secret as a GPG-encrypted file. It works across R sessions,
Docker builds, and shell scripts — making it the most unified approach.

**Why `pass` is recommended:**

- Tokens are GPG-encrypted at rest (not plaintext)
- Works in R (via `.Rprofile`), Docker builds (via build script), and CI
  (GitHub secrets handle CI separately)
- `gpg-agent` caches your passphrase so you only unlock once per session
- No extra R packages required

**First-time setup:**

```bash
# 1. Install pass (if not already installed)
# Ubuntu/Debian: sudo apt install pass
# macOS: brew install pass
# RHEL/Rocky: sudo dnf install pass

# 2. Initialise the password store (if you haven't already)
# pass init "your-gpg-key-id"

# 3. Store your Zenodo tokens
pass insert zenodo/production-token
# Enter your production token when prompted

pass insert zenodo/sandbox-token
# Enter your sandbox token when prompted

# 4. Verify
pass show zenodo/production-token
```

**Auto-loading in R sessions:**

Add this to your `~/.Rprofile` so `Sys.getenv("ZENODO_TOKEN")` and
`Sys.getenv("ZENODO_SANDBOX_TOKEN")` work automatically:

```r
# Load Zenodo tokens from pass (GPG-encrypted) if available
local({
  if (nzchar(Sys.which("pass"))) {
    load_pass <- function(env_var, pass_key) {
      token <- tryCatch(
        system2("pass", c("show", pass_key), stdout = TRUE, stderr = FALSE),
        error = function(e) character(0)
      )
      if (length(token) == 1L && nzchar(token)) {
        args <- list(token)
        names(args) <- env_var
        do.call(Sys.setenv, args)
      }
    }
    load_pass("ZENODO_TOKEN", "zenodo/production-token")
    load_pass("ZENODO_SANDBOX_TOKEN", "zenodo/sandbox-token")
  }
})
```

Then in your R code, just use `Sys.getenv()` — no manual token handling:

```r
IMPACTncd$zenodo_connect(
  token   = Sys.getenv("ZENODO_SANDBOX_TOKEN"),
  sandbox = TRUE
)
```

**Docker builds:**

The `docker_build_push.sh` script in the `docker_setup/` directory bundles the
Zenodo-managed data files that are **already present locally** (downloaded
beforehand via `zenodo_download_*()`, and listed in the inputs manifest) into
the image build context. It does **not** download from Zenodo nor read a
Zenodo token at build time — so make sure the data is downloaded locally first.

```bash
# Data must already be present locally (run zenodo_download_all() first)
cd docker_setup
./docker_build_push.sh Dockerfile.IMPACTncdENGL
```

**Manage your stored tokens:**

```bash
# List Zenodo entries
pass ls zenodo/

# Update a token
pass edit zenodo/production-token

# Delete a token
pass rm zenodo/sandbox-token
```

#### Option B: `.Renviron` File (Simple Alternative)

If you prefer not to use `pass`, the `.Renviron` file stores environment
variables that R loads at startup. Tokens are stored in **plaintext**
but the file is protected by filesystem permissions.

```r
# Open .Renviron for editing
usethis::edit_r_environ()
```

Add these lines to the file:

```
ZENODO_TOKEN=your_production_token_here
ZENODO_SANDBOX_TOKEN=your_sandbox_token_here
```

Save and restart R. The tokens are now available via
`Sys.getenv("ZENODO_TOKEN")` and `Sys.getenv("ZENODO_SANDBOX_TOKEN")`.

**Security considerations:**

- The `.Renviron` file is stored in your home directory (`~/.Renviron`)
- It is **not** in your project folder, so it won't be accidentally committed
  to git
- Set restrictive permissions: `chmod 600 ~/.Renviron` (Linux/macOS)
- Tokens are stored in **plaintext** — use `pass` (Option A) if you need
  encryption at rest

#### Option C: `keyring` R Package (Best for Windows or GUI Desktops)

On desktops with a graphical environment, `keyring` uses the operating
system's secure credential store (Windows Credential Manager, macOS Keychain,
or Linux Secret Service):

```r
# Store the token (you'll be prompted to enter it securely)
keyring::key_set("zenodo", "token")

# Retrieve the token in your scripts
token <- keyring::key_get("zenodo", "token")
IMPACTncd$zenodo_connect(token = token)
```

On servers without a desktop environment, use the **file-based backend**:

```r
kb <- keyring::backend_file$new()
kb$keyring_create("zenodo_keyring")
kb$set("zenodo_sandbox", username = "token", keyring = "zenodo_keyring")
kb$set("zenodo", username = "token", keyring = "zenodo_keyring")

# Retrieve
token <- kb$get("zenodo_sandbox", username = "token", keyring = "zenodo_keyring")
```

#### Security Best Practices

| Do | Don't |
|-----|--------|
| Use `pass` or `keyring` for encrypted storage | Store tokens in plaintext `.env` files |
| Use `.Renviron` with `chmod 600` if not using `pass` | Hardcode tokens in scripts |
| Use separate tokens for sandbox vs production | Share tokens between users |
| Regenerate tokens periodically | Use tokens with excessive permissions |
| Use GitHub secrets for CI workflows | Commit tokens to git |

## Quick Start

### Downloading Input Data (New Team Member)

If you're setting up the model for the first time, download the input data
from Zenodo. The published data is **public**, so downloading needs **no
Zenodo account or token** — just connect and download:

```{r download-quickstart}
library(IMPACTncdEngland)

# Create simulation object (skeleton mode if data is not present yet)
IMPACTncd <- Simulation$new("./inputs/sim_design.yaml")

# Connect (no token needed for downloading published data); the production
# concept DOI 10.5281/zenodo.20812409 is the default.
IMPACTncd$zenodo_connect()

# Download everything needed to run the model (inputs + PARFs + RR tables)
IMPACTncd$zenodo_download_all()
```

For testing the pipeline against the sandbox instead, pass `sandbox = TRUE`
(see [Sandbox vs Production](#sandbox-vs-production)).

### Uploading Input Data (Data Manager)

To share your input data via Zenodo:

```{r upload-quickstart}
# Connect to sandbox for testing
IMPACTncd$zenodo_connect(
  token   = Sys.getenv("ZENODO_SANDBOX_TOKEN"),
  sandbox = TRUE,
  concept_doi = NULL  # no existing record yet
)

# Upload everything (inputs + simulation outputs) in one call
IMPACTncd$zenodo_upload_all(
  title       = "IMPACTncd England Input Data",
  description = "Input and simulation data for IMPACTncd England",
  creators    = list(
    list(firstname = "Chris", lastname = "Kypridemos",
         orcid = "0000-0002-0746-9229")
  ),
  version = "1.0.0"
)

# Verify upload
IMPACTncd$zenodo_list_files()

# When satisfied, publish (IRREVERSIBLE!)
IMPACTncd$zenodo_publish()
```

## Detailed Usage

### Connecting to Zenodo

The `zenodo_connect()` method initializes the Zenodo connection and sets the
concept DOI. When `sandbox = TRUE`, it automatically defaults to the project's
sandbox concept DOI (`10.5072/zenodo.442996`):

```{r connect-details}
# Sandbox (for testing) — default concept DOI is set automatically
IMPACTncd$zenodo_connect(
  token   = Sys.getenv("ZENODO_SANDBOX_TOKEN"),
  sandbox = TRUE
)

# Production — specify your real concept DOI (no sandbox flag)
# IMPACTncd$zenodo_connect(
#   token       = Sys.getenv("ZENODO_TOKEN"),
#   concept_doi = "10.5281/zenodo.20812409"  # replace with your real DOI
# )

# Archives are created in tempdir() and automatically deleted after upload.
# To keep archives for inspection, set a custom directory:
# IMPACTncd$zenodo_connect(sandbox = TRUE, archive_dir = "./my_archives")

# Disable progress bars (useful for non-interactive sessions)
IMPACTncd$zenodo_connect(sandbox = TRUE, progress = FALSE)
```

### Uploading Input Data

The `zenodo_upload_inputs()` method creates zip archives from your input
directories, creates a Zenodo record (or new version), and uploads them:

```{r upload-inputs}
# First-time upload (creates new record)
IMPACTncd$zenodo_upload_inputs(
  title       = "IMPACTncd England Input Data",
  description = "Exposure distributions, disease burden, mortality, etc.",
  creators    = list(
    list(firstname = "Chris", lastname = "Kypridemos",
         orcid = "0000-0002-0746-9229")
  ),
  version  = "1.0.0",
  keywords = c("microsimulation", "health", "IMPACTncd", "England"),
  license  = "cc-by-sa-4.0"
)

# Upload specific directories only
IMPACTncd$zenodo_upload_inputs(
  directories = c("mortality", "exposure_distributions"),
  version     = "1.0.0",
  title       = "IMPACTncd England Input Data",
  description = "Selected input directories",
  creators    = list(
    list(firstname = "Chris", lastname = "Kypridemos")
  )
)
```

### Uploading Simulation Outputs

After uploading inputs, you can add simulation-generated files (PARFs and
compiled RR tables) to the **same** Zenodo record:

```{r upload-simulation}
# Upload simulation outputs (parf/ and rr/ directories)
IMPACTncd$zenodo_upload_simulation()

# With custom directories or exclusion patterns
IMPACTncd$zenodo_upload_simulation(
  simulation_base = "./simulation",
  directories     = c("parf", "rr"),
  exclude_file_patterns = c("\\.R$", "\\.csv$")
)
```

### Uploading Everything at Once

The `zenodo_upload_all()` convenience method combines input and simulation
uploads in a single call:

```{r upload-all}
IMPACTncd$zenodo_upload_all(
  title       = "IMPACTncd England Input Data",
  description = "Complete input and simulation data",
  creators    = list(
    list(firstname = "Chris", lastname = "Kypridemos",
         orcid = "0000-0002-0746-9229")
  ),
  version = "1.0.0"
)
```

### Creating New Versions

When your input data changes, create a new version of the existing record:

```{r new-version}
# Connect with existing concept DOI
IMPACTncd$zenodo_connect(
  token   = Sys.getenv("ZENODO_SANDBOX_TOKEN"),
  sandbox = TRUE  # concept_doi defaults to "10.5072/zenodo.442996"
)

# Upload as new version
IMPACTncd$zenodo_upload_inputs(version = "1.1.0")

# Or upload everything as new version
IMPACTncd$zenodo_upload_all(version = "1.1.0")

# Publish
IMPACTncd$zenodo_publish()
```

### Listing Remote Files

After uploading, verify what's on the server:

```{r list-files}
files <- IMPACTncd$zenodo_list_files()
print(files)
```

This also works with draft records (where zen4R's built-in listing may
return empty results). The method automatically falls back to a direct API
query.

### Downloading Input Data

Downloads zip archives from the Zenodo record and extracts them into
`./inputs/`. By default, directories that already exist locally are
**skipped** — pass `overwrite = TRUE` to replace them.

```{r download-details}
# Check what's available vs what you have locally
status <- IMPACTncd$zenodo_check_inputs()
print(status)

# Download missing directories (skips existing ones)
IMPACTncd$zenodo_download_inputs()

# Download specific directories
IMPACTncd$zenodo_download_inputs(
  directories = c("mortality", "exposure_distributions")
)

# Force re-download (overwrites local directories)
IMPACTncd$zenodo_download_inputs(overwrite = TRUE)
```

### Downloading PARF and RR Data

If PARFs and compiled RR tables were uploaded to the same Zenodo record
(via `zenodo_upload_simulation()` or `zenodo_upload_all()`), you can
download them separately:

```{r download-simulation}
# Download pre-computed PARFs and RR tables
IMPACTncd$zenodo_download_PARFs_RRs()

# Download only PARF files (skip RR)
IMPACTncd$zenodo_download_PARFs_RRs(directories = "parf")

# Force re-download (overwrites existing)
IMPACTncd$zenodo_download_PARFs_RRs(overwrite = TRUE)
```

This is useful when:

- A colleague has already generated PARFs/RR tables and you want to reuse
  them instead of re-running the generation step (which can take hours)
- You're setting up a new machine and want to skip time-consuming
  computation

### Downloading Everything at Once

The `zenodo_download_all()` convenience method downloads both input data
**and** simulation outputs in a single call:

```{r download-all}
# New team member — connect and download everything
IMPACTncd$zenodo_connect(
  token   = Sys.getenv("ZENODO_SANDBOX_TOKEN"),
  sandbox = TRUE  # concept_doi defaults to "10.5072/zenodo.442996"
)
IMPACTncd$zenodo_download_all()

# Force re-download everything
IMPACTncd$zenodo_download_all(overwrite = TRUE)
```

This calls `zenodo_download_inputs()` followed by
`zenodo_download_PARFs_RRs()` internally. Input data goes to `./inputs/`
and simulation data goes to `./simulation/`.

## Advanced: Archive Control

### Exclude Patterns

Control which directories and files are included in archives:

```{r exclude-patterns}
IMPACTncd$zenodo_upload_inputs(
  version = "1.0.0",
  title   = "...", description = "...", creators = list(...),

  # Skip entire directories matching these regex patterns
  exclude_patterns = c(
    "^unprocessed$",  # Exact match: skip "unprocessed" directory
    "_backup$",       # Ends with "_backup"
    "_old$",          # Ends with "_old"
    "scripts$",       # Ends with "scripts"
    "validation$",    # Ends with "validation"
    "^RR$"            # Source RR .csvy files — tracked in git, not Zenodo
  ),

  # Skip individual files matching these regex patterns
  exclude_file_patterns = c(
    "\\.R$",     # R scripts
    "\\.Rmd$",   # R Markdown files
    "\\.yaml$",  # YAML config files
    "\\.html$"   # HTML reports
  )
)
```

The default exclude patterns are designed for the IMPACTncd project:

- **Directory patterns**: `^unprocessed$`, `_backup$`, `_old$`, `scripts$`,
  `validation$`, `^RR$`
- **Input file patterns**: `\\.R$`, `\\.Rmd$`
- **Simulation file patterns**: `\\.R$`, `\\.csv$`

`^RR$` excludes the `inputs/RR/` directory because the source `.csvy`
relative-risk files are **tracked in git** (see `.gitignore` allow-list
`!inputs/RR/**/*.csvy`), not uploaded to Zenodo. The compiled RR `.fst`
files in `simulation/rr/` are still uploaded to Zenodo via
`zenodo_upload_simulation()`.

### Group-By-Prefix

When `group_by_prefix = TRUE` (the default), subdirectories sharing a common
prefix are combined into a single archive. This reduces the number of
individual uploads:

```
inputs/disease_burden/
  ├── cancer_2020/    ─┐
  ├── cancer_2021/    ─┤── disease_burden_cancer.zip
  ├── cancer_2022/    ─┘
  ├── chd_2020/       ─┐
  ├── chd_2021/       ─┤── disease_burden_chd.zip
  ├── chd_2022/       ─┘
  └── stroke_2020/    ─── disease_burden_stroke.zip
```

Set `group_by_prefix = FALSE` to create one archive per subdirectory instead.
Note that Zenodo enforces a per-record file limit (100 files at the time of
writing — an external Zenodo policy, not checked by this package beyond a
pre-upload guard in the upload methods), so grouping by prefix is recommended
when a directory has many subdirectories. The upload methods will error before
uploading if the archive count for a record would exceed 100.

### Compression Level

The `compression_level` parameter controls the zip compression level (0-9):

| Level | Speed | Size | Use case |
|-------|-------|------|----------|
| 0 | Fastest | Largest | Already-compressed files (qs, fst) |
| 6 | Balanced | Medium | **Default** — good for most data |
| 9 | Slowest | Smallest | When bandwidth is limited |

```{r compression}
# Fast archiving for pre-compressed files
IMPACTncd$zenodo_upload_inputs(
  ...,
  compression_level = 0  # store only, no compression
)
```

### Parallel Archive Creation

Archives are created in parallel by default using PSOCK clusters (which are
safe on all platforms, unlike forking). The number of cores defaults to
`clusternumber` from your simulation design YAML:

```{r parallel}
# Use specific number of cores
IMPACTncd$zenodo_upload_inputs(
  ...,
  multicore = TRUE,
  n_cores   = 10L
)

# Disable parallelism (single-threaded)
IMPACTncd$zenodo_upload_inputs(
  ...,
  multicore = FALSE
)
```

### Preview Archives Before Uploading

Create archives without uploading to inspect sizes and contents:

```{r preview-archives}
# Create archives locally (no Zenodo connection needed)
archives <- IMPACTncd$zenodo_create_archives()

# Inspect results
print(archives[, .(
  archive    = basename(archive_path),
  size_mb    = round(size_bytes / 1024^2, 1),
  file_count = file_count
)])
```

### Automatic Git Tracking Management

When `update_gitignore = TRUE` (the default for upload methods), two things
happen automatically:

1. **`.gitignore` update** — Archived data files are added to `.gitignore`
   using efficient directory/extension patterns (e.g., `inputs/mortality/*.fst`).
   This prevents new data files from being accidentally committed in the future.

2. **Git index cleanup** — Any data files that were *already tracked* by git
   are removed from the git index via `git rm --cached`. This is critical
   because `.gitignore` only prevents *new* untracked files from being staged;
   files that were previously committed continue to be tracked regardless of
   `.gitignore` rules. The `--cached` flag means files are **kept on disk** —
   only the git tracking is removed.

After uploading to Zenodo, your data files live in two places: on your local
disk (for the model to use) and on Zenodo (for reproducibility and sharing).
Anyone cloning the repository will need to run `zenodo_download_inputs()` or
`zenodo_download_PARFs_RRs()` to get the data files.

## Input Directory Structure

The pipeline archives each top-level directory in `./inputs/` separately:

```
inputs/
├── exposure_distributions/    -> exposure_distributions.zip
│   ├── bmi/
│   ├── smoking/
│   └── ...
├── disease_burden/            -> disease_burden_*.zip (grouped by prefix)
│   ├── chd_incd/
│   ├── stroke_prvl/
│   └── ...
├── mortality/                 -> mortality.zip
├── RR/                        (NOT archived — tracked in git instead;
│   ├── bmi~chd.csvy            see "Two RR locations" note below)
│   └── ...
├── pop_estimates_lsoa/        -> pop_estimates_lsoa.zip
├── pop_projections/           -> pop_projections.zip
└── aux_data/                  -> aux_data.zip

simulation/
├── parf/                      -> parf_*.zip (grouped by disease prefix)
│   ├── PARF_af_<hash>.qs
│   ├── PARF_chd_<hash>.qs
│   └── ...
└── rr/                        -> rr.zip
    ├── chd.fst
    ├── stroke.fst
    └── ...
```

Each archive contains the full directory structure, so extraction recreates
the correct paths.

### Two RR locations

The model uses **two distinct RR directories** with different storage
strategies:

| Location | Contents | Storage |
|----------|----------|---------|
| `inputs/RR/` | Source CSVY files (e.g. `bmi~chd.csvy`) — small, human-readable | **Git** (allow-listed in `.gitignore`) |
| `simulation/rr/` | Compiled `.fst` files generated from the CSVYs — large, binary | **Zenodo** (via `zenodo_upload_simulation()`) |

This is why `^RR$` is in the default `exclude_patterns` for input uploads:
the source files belong in version control, not Zenodo.

## Sandbox vs Production

Zenodo operates two **completely separate systems**. They do not share
accounts, tokens, data, or DOIs. Understanding this distinction is essential
before uploading or downloading data.

| | Sandbox | Production |
|---|---|---|
| **Purpose** | Development and testing | Permanent archiving and citation |
| **Web UI** | [sandbox.zenodo.org](https://sandbox.zenodo.org/) | [zenodo.org](https://zenodo.org/) |
| **Account** | Separate registration required | Separate registration required |
| **Token env var** | `ZENODO_SANDBOX_TOKEN` | `ZENODO_TOKEN` |
| **DOI prefix** | `10.5072/zenodo.XXXXXX` | `10.5281/zenodo.XXXXXX` |
| **Data permanence** | Periodically wiped — records may disappear | Permanent — published records cannot be deleted |
| **Publishable** | Yes, but DOIs do not resolve publicly | Yes — DOIs are globally resolvable and citable |
| **R flag** | `sandbox = TRUE` | `sandbox = FALSE` (default) |

### DOI Prefixes

The DOI prefix tells you instantly which system a record belongs to:

- **`10.5072/zenodo.442996`** — This is a **sandbox** DOI. It only works
  with `sandbox = TRUE` and the sandbox token.
- **`10.5281/zenodo.20812409`** — This is a **production** DOI. It only
  works with `sandbox = FALSE` (the default) and the production token.

A sandbox DOI will **not** resolve on production Zenodo and vice versa.
Using a sandbox DOI without `sandbox = TRUE` will give a "Record not
found" error.

### When to Use Each

**Use sandbox** when you are:

- Testing the upload/download pipeline for the first time
- Experimenting with archive settings (compression, grouping, exclusions)
- Developing or debugging Zenodo integration code
- Running CI tests that exercise the Zenodo workflow

**Use production** when you are:

- Sharing input data with collaborators who need to run the model
- Publishing a version of the data alongside a paper or report
- Setting up a reproducible workflow that must persist long-term

### Connecting to Sandbox

When `sandbox = TRUE`, the project's sandbox concept DOI
(`10.5072/zenodo.442996`) is set automatically:

```{r sandbox}
# Sandbox — concept DOI defaults to 10.5072/zenodo.442996
IMPACTncd$zenodo_connect(
  token   = Sys.getenv("ZENODO_SANDBOX_TOKEN"),
  sandbox = TRUE
)

# Override the default sandbox DOI if needed
IMPACTncd$zenodo_connect(
  token       = Sys.getenv("ZENODO_SANDBOX_TOKEN"),
  sandbox     = TRUE,
  concept_doi = "10.5072/zenodo.999999"
)
```

### Connecting to Production

Production is the default. `zenodo_connect()` with no arguments uses the
published IMPACTncd England input-data concept DOI
(`10.5281/zenodo.20812409`) and connects anonymously (no token needed to
download published data):

```{r production-connect}
# Simplest: defaults to the published production record, no token
IMPACTncd$zenodo_connect()

# Equivalent, being explicit (and supplying a token if you also want to
# upload/publish):
IMPACTncd$zenodo_connect(
  token       = Sys.getenv("ZENODO_TOKEN"),
  concept_doi = "10.5281/zenodo.20812409"
)
```

### Transitioning from Sandbox to Production

Once you have validated your upload/download workflow on sandbox, follow
these steps to move to production:

**1. Create a production Zenodo account and token**

Go to [zenodo.org](https://zenodo.org/account/settings/applications/tokens/new/)
and create a personal access token with `deposit:write` and
`deposit:actions` scopes. Store it as `ZENODO_TOKEN` (see
[Secure Token Management](#secure-token-management) above).

**2. Upload to production (first time — creates a new record)**

```{r transition-upload}
IMPACTncd$zenodo_connect(
  token       = Sys.getenv("ZENODO_TOKEN"),
  concept_doi = NULL  # no existing production record yet
)

IMPACTncd$zenodo_upload_all(
  title       = "IMPACTncd England Input Data",
  description = "Input and simulation data for IMPACTncd England",
  creators    = list(
    list(firstname = "Chris", lastname = "Kypridemos",
         orcid = "0000-0002-0746-9229")
  ),
  version = "1.0.0"
)
```

**3. Note your new production concept DOI**

After uploading, the record object contains your new concept DOI. Save it
— you will use it for all future connections and share it with
collaborators:

```{r transition-doi}
# The concept DOI is reported in the create_new_record/publish log messages.
# You can also see all versions and their DOIs:
IMPACTncd$zenodo_get_versions()
```

**4. Verify and publish**

Review the draft on the Zenodo website, then publish. **Publishing is
irreversible** — the DOI becomes permanent and the files cannot be
modified (you can only create new versions).

```{r transition-publish}
IMPACTncd$zenodo_list_files()
IMPACTncd$zenodo_publish()
```

**5. Share with collaborators**

Collaborators now download from production using your concept DOI:

```{r transition-download}
IMPACTncd$zenodo_connect(
  token       = Sys.getenv("ZENODO_TOKEN"),
  concept_doi = "10.5281/zenodo.20812409"  # replace with your real DOI
)
IMPACTncd$zenodo_download_all()
```

The concept DOI always resolves to the latest published version, so
collaborators automatically get the most recent data when they download.

## The ZenodoAssetManager Class

Under the hood, the Simulation class delegates to a `ZenodoAssetManager`
R6 class. You can use it directly for more control:

```{r direct-manager}
library(IMPACTncdEngland)

# Create manager directly
zam <- ZenodoAssetManager$new(
  hash_file = "./simulation/zenodo_manifest.csv",
  logs      = TRUE,
  sandbox   = TRUE
)

# Connect
zam$connect(token = Sys.getenv("ZENODO_SANDBOX_TOKEN"))
zam$set_progress_callback(upload = TRUE, download = TRUE)

# Set concept DOI
zam$set_concept_doi("10.5072/zenodo.442996")

# Get record info
zam$get_record()
print(zam$record)

# List remote files
files <- zam$list_remote_files()
print(files)

# Create archives with full control
archives <- zam$create_input_archives(
  input_base            = "./inputs",
  directories           = c("mortality", "exposure_distributions"),
  exclude_patterns      = c("^unprocessed$"),
  exclude_file_patterns = c("\\.R$", "\\.Rmd$"),
  group_by_prefix       = TRUE,
  compression_level     = 6L,
  multicore             = TRUE,
  n_cores               = 10L,
  update_gitignore      = TRUE
)

# Upload individual archives
for (i in seq_len(nrow(archives))) {
  zam$upload_archive(archives$archive_path[i])
}
```

## Hash Manifests

The pipeline uses file hashes (xxhash64) to detect changes:

```{r manifests}
# Create a manifest of all input files
manifest <- IMPACTncd$zenodo_create_manifest()
print(manifest)

# The ZenodoAssetManager also provides lower-level manifest operations:
zam <- ZenodoAssetManager$new()

manifest <- zam$compute_manifest("./inputs", recursive = TRUE)
zam$save_manifest(manifest)

old_manifest <- zam$load_manifest()
diff <- zam$compare_manifests(manifest, old_manifest)
print(diff$modified)  # Files that changed
print(diff$added)     # New files
print(diff$removed)   # Deleted files
```

## Troubleshooting

### Common Issues

**1. "Not connected to Zenodo"**

```r
# Solution: Call zenodo_connect() first
IMPACTncd$zenodo_connect(sandbox = TRUE)
```

**2. "... requires a Zenodo token, but the manager is connected anonymously"**

Downloading works anonymously, but uploading/publishing needs a token.
Reconnect with a token:

```r
# Solution: Set environment variable or pass token directly
Sys.setenv(ZENODO_TOKEN = "your_token")
IMPACTncd$zenodo_connect(token = Sys.getenv("ZENODO_TOKEN"))
# (use ZENODO_SANDBOX_TOKEN and sandbox = TRUE for the sandbox)
```

**3. "Record not found for DOI"**

```r
# Solution: Verify DOI format (should be the concept DOI, not a URL)
# Correct: "10.5072/zenodo.442996"
# Wrong:   "https://doi.org/10.5072/zenodo.442996"
IMPACTncd$zenodo_set_doi("10.5072/zenodo.442996")
```

**4. HTTP 403 Forbidden on file listing**

This can happen when zen4R's internal URL does not match the Zenodo server.
The `list_remote_files()` method automatically falls back to a direct API
query. If you still get 403, ensure your token has the correct scopes
(`deposit:write`, `deposit:actions`).

**5. Empty file listing after upload**

zen4R's `listFiles(pretty = TRUE)` can return empty results for **draft**
records on the new Zenodo InvenioRDM backend. The `zenodo_list_files()` method
handles this automatically by falling back to a direct API query to
`/api/records/{id}/draft/files`.

**6. "Missing publisher field required for DOI registration"**

This error occurs when publishing. The `publisher` metadata field is required
by Zenodo's InvenioRDM backend for DOI registration. It defaults to
`"Zenodo"` in `zenodo_upload_inputs()`, so this should only happen if you
created the record manually via the ZenodoAssetManager.

```r
# Fix: Set publisher on the record before publishing
manager$record$setPublisher("Zenodo")
```

**7. Upload shows progress but fails silently**

zen4R may show "No bucket link" warnings and `x` markers during upload, but
the files may actually be uploaded successfully. Always verify with
`zenodo_list_files()` after uploading.

### Network Issues

For slow connections or large files:

```r
# Download specific directories first (start with smaller ones)
IMPACTncd$zenodo_download_inputs(
  directories = c("mortality", "RR")
)
```

## Complete Example Workflow

### Step 1: Test on Sandbox

Before committing to permanent production records, validate your workflow
on sandbox:

```{r workflow-sandbox-test}
library(IMPACTncdEngland)

# 1. Create simulation object
IMPACTncd <- Simulation$new("./inputs/sim_design.yaml")

# 2. Connect to sandbox (no existing record)
IMPACTncd$zenodo_connect(
  token       = Sys.getenv("ZENODO_SANDBOX_TOKEN"),
  sandbox     = TRUE,
  concept_doi = NULL  # creates a new sandbox record
)

# 3. (Optional) Preview archives first
archives <- IMPACTncd$zenodo_create_archives()
print(archives[, .(archive = basename(archive_path),
                    size_mb = round(size_bytes / 1024^2, 1))])

# 4. Upload everything
IMPACTncd$zenodo_upload_all(
  title       = "IMPACTncd England Input Data",
  description = paste(
    "Input and simulation data for the IMPACTncd England",
    "microsimulation model, including exposure distributions,",
    "disease burden estimates, mortality data, PARFs, and",
    "compiled relative risk tables."
  ),
  creators = list(
    list(firstname = "Chris", lastname = "Kypridemos",
         orcid = "0000-0002-0746-9229")
  ),
  version = "1.0.0"
)

# 5. Verify — check file count and sizes look correct
files <- IMPACTncd$zenodo_list_files()
cat("Uploaded", nrow(files), "files\n")

# 6. Publish sandbox record (safe — sandbox records are not permanent)
IMPACTncd$zenodo_publish()

# 7. Test download in a clean directory to confirm the round-trip works
IMPACTncd$zenodo_download_all(overwrite = TRUE)
```

### Step 2: Publish to Production

Once you're confident the pipeline works, repeat on production. This
creates a **permanent, citable record** — the DOI cannot be retracted.

```{r workflow-production}
# Connect to production (no existing record)
IMPACTncd$zenodo_connect(
  token       = Sys.getenv("ZENODO_TOKEN"),
  concept_doi = NULL  # first upload — creates a new production record
)

# Upload (same parameters as sandbox)
IMPACTncd$zenodo_upload_all(
  title       = "IMPACTncd England Input Data",
  description = paste(
    "Input and simulation data for the IMPACTncd England",
    "microsimulation model, including exposure distributions,",
    "disease burden estimates, mortality data, PARFs, and",
    "compiled relative risk tables."
  ),
  creators = list(
    list(firstname = "Chris", lastname = "Kypridemos",
         orcid = "0000-0002-0746-9229")
  ),
  version = "1.0.0"
)

# Verify upload
files <- IMPACTncd$zenodo_list_files()
cat("Uploaded", nrow(files), "files\n")

# IMPORTANT: Save your concept DOI before publishing!
# You will need it for all future connections.
# The concept DOI is visible on the Zenodo web UI for the draft record,
# or in: IMPACTncd$zenodo_get_versions()

# Publish (IRREVERSIBLE — the DOI becomes permanent)
IMPACTncd$zenodo_publish()
```

### New Team Member (Downloading Data)

Share your **production concept DOI** with collaborators. They download
with:

```{r workflow-download}
library(IMPACTncdEngland)
IMPACTncd <- Simulation$new("./inputs/sim_design.yaml")

# Connect to production with the project's concept DOI
IMPACTncd$zenodo_connect(
  token       = Sys.getenv("ZENODO_TOKEN"),
  concept_doi = "10.5281/zenodo.20812409"  # replace with real concept DOI
)

# Check what's available
status <- IMPACTncd$zenodo_check_inputs()

# Download everything (inputs + pre-computed PARFs + RR tables)
IMPACTncd$zenodo_download_all()

# Or download only inputs (if you want to generate PARFs/RR yourself)
# IMPACTncd$zenodo_download_inputs()

# Run simulation
IMPACTncd$run(mc = 1:10, multicore = TRUE, scenario_nam = "sc0")
```

### Publishing Updated Data (New Version)

When input data changes, create a new version of the **existing**
production record:

```{r workflow-update}
# Connect with existing production concept DOI
IMPACTncd$zenodo_connect(
  token       = Sys.getenv("ZENODO_TOKEN"),
  concept_doi = "10.5281/zenodo.20812409"  # your production concept DOI
)

# Upload as new version (creates new version of existing record)
IMPACTncd$zenodo_upload_all(version = "1.1.0")

# Verify and publish
IMPACTncd$zenodo_list_files()
IMPACTncd$zenodo_publish()

# Check version history
IMPACTncd$zenodo_get_versions()
```

## API Reference

### Simulation Class Methods

| Method | Description |
|--------|-------------|
| `zenodo_connect(token, concept_doi, sandbox, archive_dir, progress)` | Connect to Zenodo and set concept DOI |
| `zenodo_set_doi(concept_doi)` | Change the concept DOI after connecting |
| `zenodo_check_inputs(input_base, directories)` | Check local vs remote sync status |
| `zenodo_download_inputs(input_base, directories, overwrite)` | Download and extract input archives |
| `zenodo_download_PARFs_RRs(simulation_base, directories, overwrite)` | Download and extract simulation archives |
| `zenodo_download_all(input_base, ..., simulation_base, ..., overwrite)` | Download inputs + simulation in one call |
| `zenodo_upload_inputs(input_base, ..., publish)` | Create archives and upload inputs |
| `zenodo_upload_simulation(simulation_base, directories, ...)` | Upload simulation outputs (PARF, RR) |
| `zenodo_upload_all(input_base, simulation_base, ..., publish)` | Upload inputs + simulation in one call |
| `zenodo_publish()` | Publish the draft record (**irreversible!**) |
| `zenodo_list_files(concept_doi)` | List all files in the current record |
| `zenodo_get_versions()` | Get version history with DOIs |
| `zenodo_create_archives(input_base, ...)` | Create archives locally (no upload) |
| `zenodo_create_manifest(input_base, save)` | Create file hash manifest |

### Archive Creation Parameters

These parameters are available in `zenodo_upload_inputs()`,
`zenodo_upload_all()`, and `zenodo_create_archives()`:

| Parameter | Default | Description |
|-----------|---------|-------------|
| `exclude_patterns` | `c("^unprocessed$", "_backup$", "_old$", "scripts$", "validation$", "^RR$")` | Regex patterns for directories to skip |
| `exclude_file_patterns` | `c("\\.R$", "\\.Rmd$")` | Regex patterns for files to skip |
| `group_by_prefix` | `TRUE` | Combine subdirectories by shared prefix |
| `compression_level` | `6` | Zip compression (0=store, 9=max) |
| `multicore` | `TRUE` | Parallel archive creation |
| `n_cores` | `clusternumber` from YAML | Number of CPU cores |
| `update_gitignore` | `TRUE` | Auto-update .gitignore and remove tracked data files from git index |

## References

- [Zenodo](https://zenodo.org/) - Open repository hosted by CERN
- [zen4R package](https://github.com/eblondel/zen4R) - R interface to Zenodo
  API
- [Zenodo REST API](https://developers.zenodo.org/) - API documentation
- [DOI Foundation](https://www.doi.org/) - About Digital Object Identifiers
