---
title: "Managing Input Data with Zenodo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Managing Input Data with Zenodo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

IMPACTncdEngland includes a robust pipeline for managing input data files
via [Zenodo](https://zenodo.org/), a general-purpose open repository hosted
by CERN. This enables:

- **Reproducibility**: DOI-based versioning ensures you always know which data
  version was used
- **Collaboration**: Easy sharing of input datasets across team members
- **Archiving**: Long-term preservation of input data with persistent
  identifiers
- **Version Control**: Track changes to input data over time

The pipeline uses the `zen4R` package to interact with the Zenodo REST API,
and supports **parquet**, **fst**, and **qs** file formats commonly used in
the simulation. It can archive both **input data** (exposure distributions,
disease burden, mortality, etc.) and **simulation outputs** (PARFs, compiled
RR tables).

### What is a DOI?

A **Digital Object Identifier (DOI)** is a permanent, unique identifier for a
digital resource. When you publish data on Zenodo, it receives a DOI that will
always resolve to that exact version of the data, even decades later. This is
the standard way to cite research data in academic publications.

Zenodo provides two types of DOIs:

| Type | Example | Resolves to |
|------|---------|-------------|
| **Concept DOI** | `10.5281/zenodo.12345` | Always the *latest* published version |
| **Version DOI** | `10.5281/zenodo.12346` | A *specific* version (immutable) |

Use the **concept DOI** in your code and share it with collaborators. Cite the
**version DOI** in publications for exact reproducibility.

## Prerequisites

### 1. Install Required Packages

The following packages are required (they are Suggests in the package):

```{r install-packages}
install.packages(c("zen4R", "zip", "digest", "httr2"))
# Optional: for secure token storage
install.packages("keyring")
```

### 2. Get a Zenodo Access Token

1. Create a Zenodo account at [zenodo.org](https://zenodo.org/)
2. Go to
   [Applications > Personal access tokens](https://zenodo.org/account/settings/applications/tokens/new/)
3. Create a new token with **deposit:write** and **deposit:actions** scopes
4. Save the token securely (see below)

**For testing**, create a separate sandbox account and token at
[sandbox.zenodo.org](https://sandbox.zenodo.org/account/settings/applications/tokens/new/).
Sandbox records are not permanent and do not count as real publications.

### 3. Secure Token Management

Your Zenodo token grants full access to your Zenodo deposits. **Never commit
tokens to version control!**

#### Option A: `.Renviron` File (Recommended for Personal Use)

The `.Renviron` file stores environment variables that R loads at startup:

```r
# Open .Renviron for editing
usethis::edit_r_environ()
```

Add these lines to the file:

```
ZENODO_TOKEN=your_production_token_here
ZENODO_SANDBOX_TOKEN=your_sandbox_token_here
```

Save and restart R. The tokens are now available via
`Sys.getenv("ZENODO_TOKEN")` and `Sys.getenv("ZENODO_SANDBOX_TOKEN")`.

**Security considerations:**

- The `.Renviron` file is stored in your home directory (`~/.Renviron`)
- It is **not** in your project folder, so it won't be accidentally committed
  to git
- Set restrictive permissions: `chmod 600 ~/.Renviron` (Linux/macOS)

#### Option B: File-Based Keyring (Best for Servers)

On servers without a desktop environment (where GNOME Keyring / KWallet
are not available), use the `keyring` package with a **file-based backend**.
The keyring file is stored at `~/.config/r-keyring/` and encrypted with a
master password you choose.

**First-time setup** (run these commands interactively in R console):

```r
# 1. Install keyring if needed
install.packages("keyring")

# 2. Create the file backend and keyring
kb <- keyring::backend_file$new()
kb$keyring_create("zenodo_keyring")
# Enter a master password when prompted (characters won't show - that's normal)

# 3. Store your Zenodo sandbox token
kb$set("zenodo_sandbox", username = "token", keyring = "zenodo_keyring")
# Enter your Zenodo sandbox token when prompted

# 4. Store your Zenodo production token (when ready)
kb$set("zenodo", username = "token", keyring = "zenodo_keyring")
# Enter your Zenodo production token when prompted
```

**Retrieve your token in scripts:**

```r
# Helper function for file-based keyring
get_zenodo_token <- function(sandbox = TRUE,
                             keyring_name = "zenodo_keyring") {
  kb <- keyring::backend_file$new()
  service <- if (sandbox) "zenodo_sandbox" else "zenodo"
  kb$get(service, username = "token", keyring = keyring_name)
}

# Use it
token <- get_zenodo_token(sandbox = TRUE)
IMPACTncd$zenodo_connect(token = token, sandbox = TRUE)
```

**Manage your stored keys:**

```r
kb <- keyring::backend_file$new()

# List stored keys
kb$list(keyring = "zenodo_keyring")

# List available keyrings
kb$keyring_list()

# Delete a stored token
kb$delete("zenodo_sandbox", username = "token", keyring = "zenodo_keyring")
```

#### Option C: `keyring` with OS Backend (Most Secure on Desktops)

On desktops with a graphical environment, `keyring` uses the operating
system's secure credential store (Windows Credential Manager, macOS Keychain,
or Linux Secret Service):

```r
# Store the token (you'll be prompted to enter it securely)
keyring::key_set("zenodo", "token")

# Retrieve the token in your scripts
token <- keyring::key_get("zenodo", "token")
IMPACTncd$zenodo_connect(token = token)
```

#### Security Best Practices

| Do | Don't |
|-----|--------|
| Use `.Renviron` or `keyring` | Hardcode tokens in scripts |
| Add `.Renviron` to `.gitignore` if project-local | Commit tokens to git |
| Use separate tokens for sandbox vs production | Share tokens between users |
| Regenerate tokens periodically | Use tokens with excessive permissions |
| Revoke tokens you no longer need | Log or print tokens in output |

## Quick Start

### Downloading Input Data (New Team Member)

If you're setting up the model for the first time, download the input data
from Zenodo:

```{r download-quickstart}
library(IMPACTncdEngland)

# Create simulation object
IMPACTncd <- Simulation$new("./inputs/sim_design.yaml")

# Connect to Zenodo sandbox with the project's concept DOI
IMPACTncd$zenodo_connect(
  token   = Sys.getenv("ZENODO_SANDBOX_TOKEN"),
  sandbox = TRUE  # concept_doi defaults to "10.5072/zenodo.442996"
)

# Check what's available
IMPACTncd$zenodo_check_inputs()

# Download all missing inputs
IMPACTncd$zenodo_download_inputs()
```

### Uploading Input Data (Data Manager)

To share your input data via Zenodo:

```{r upload-quickstart}
# Connect to sandbox for testing
IMPACTncd$zenodo_connect(
  token   = Sys.getenv("ZENODO_SANDBOX_TOKEN"),
  sandbox = TRUE,
  concept_doi = NULL  # no existing record yet
)

# Upload everything (inputs + simulation outputs) in one call
IMPACTncd$zenodo_upload_all(
  title       = "IMPACTncd England Input Data",
  description = "Input and simulation data for IMPACTncd England",
  creators    = list(
    list(firstname = "Chris", lastname = "Kypridemos",
         orcid = "0000-0002-0746-9229")
  ),
  version = "1.0.0"
)

# Verify upload
IMPACTncd$zenodo_list_files()

# When satisfied, publish (IRREVERSIBLE!)
IMPACTncd$zenodo_publish()
```

## Detailed Usage

### Connecting to Zenodo

The `zenodo_connect()` method initializes the Zenodo connection and sets the
concept DOI. When `sandbox = TRUE`, it automatically defaults to the project's
sandbox concept DOI (`10.5072/zenodo.442996`):

```{r connect-details}
# Sandbox (for testing) ŌĆö default concept DOI is set automatically
IMPACTncd$zenodo_connect(
  token   = Sys.getenv("ZENODO_SANDBOX_TOKEN"),
  sandbox = TRUE
)

# Production ŌĆö specify your real concept DOI (no sandbox flag)
# IMPACTncd$zenodo_connect(
#   token       = Sys.getenv("ZENODO_TOKEN"),
#   concept_doi = "10.5281/zenodo.XXXXXXX"  # replace with your real DOI
# )

# Archives are created in tempdir() and automatically deleted after upload.
# To keep archives for inspection, set a custom directory:
# IMPACTncd$zenodo_connect(sandbox = TRUE, archive_dir = "./my_archives")

# Disable progress bars (useful for non-interactive sessions)
IMPACTncd$zenodo_connect(sandbox = TRUE, progress = FALSE)
```

### Uploading Input Data

The `zenodo_upload_inputs()` method creates zip archives from your input
directories, creates a Zenodo record (or new version), and uploads them:

```{r upload-inputs}
# First-time upload (creates new record)
IMPACTncd$zenodo_upload_inputs(
  title       = "IMPACTncd England Input Data",
  description = "Exposure distributions, disease burden, mortality, etc.",
  creators    = list(
    list(firstname = "Chris", lastname = "Kypridemos",
         orcid = "0000-0002-0746-9229")
  ),
  version  = "1.0.0",
  keywords = c("microsimulation", "health", "IMPACTncd", "England"),
  license  = "cc-by-sa-4.0"
)

# Upload specific directories only
IMPACTncd$zenodo_upload_inputs(
  directories = c("mortality", "exposure_distributions"),
  version     = "1.0.0",
  title       = "IMPACTncd England Input Data",
  description = "Selected input directories",
  creators    = list(
    list(firstname = "Chris", lastname = "Kypridemos")
  )
)
```

### Uploading Simulation Outputs

After uploading inputs, you can add simulation-generated files (PARFs and
compiled RR tables) to the **same** Zenodo record:

```{r upload-simulation}
# Upload simulation outputs (parf/ and rr/ directories)
IMPACTncd$zenodo_upload_simulation()

# With custom directories or exclusion patterns
IMPACTncd$zenodo_upload_simulation(
  simulation_base = "./simulation",
  directories     = c("parf", "rr"),
  exclude_file_patterns = c("\\.R$", "\\.csv$")
)
```

### Uploading Everything at Once

The `zenodo_upload_all()` convenience method combines input and simulation
uploads in a single call:

```{r upload-all}
IMPACTncd$zenodo_upload_all(
  title       = "IMPACTncd England Input Data",
  description = "Complete input and simulation data",
  creators    = list(
    list(firstname = "Chris", lastname = "Kypridemos",
         orcid = "0000-0002-0746-9229")
  ),
  version = "1.0.0"
)
```

### Creating New Versions

When your input data changes, create a new version of the existing record:

```{r new-version}
# Connect with existing concept DOI
IMPACTncd$zenodo_connect(
  token   = Sys.getenv("ZENODO_SANDBOX_TOKEN"),
  sandbox = TRUE  # concept_doi defaults to "10.5072/zenodo.442996"
)

# Upload as new version
IMPACTncd$zenodo_upload_inputs(version = "1.1.0")

# Or upload everything as new version
IMPACTncd$zenodo_upload_all(version = "1.1.0")

# Publish
IMPACTncd$zenodo_publish()
```

### Listing Remote Files

After uploading, verify what's on the server:

```{r list-files}
files <- IMPACTncd$zenodo_list_files()
print(files)
```

This also works with draft records (where zen4R's built-in listing may
return empty results). The method automatically falls back to a direct API
query.

### Downloading Input Data

Downloads zip archives from the Zenodo record and extracts them into
`./inputs/`. By default, directories that already exist locally are
**skipped** ŌĆö pass `overwrite = TRUE` to replace them.

```{r download-details}
# Check what's available vs what you have locally
status <- IMPACTncd$zenodo_check_inputs()
print(status)

# Download missing directories (skips existing ones)
IMPACTncd$zenodo_download_inputs()

# Download specific directories
IMPACTncd$zenodo_download_inputs(
  directories = c("mortality", "exposure_distributions")
)

# Force re-download (overwrites local directories)
IMPACTncd$zenodo_download_inputs(overwrite = TRUE)
```

### Downloading Simulation Data

If PARFs and compiled RR tables were uploaded to the same Zenodo record
(via `zenodo_upload_simulation()` or `zenodo_upload_all()`), you can
download them separately:

```{r download-simulation}
# Download pre-computed PARFs and RR tables
IMPACTncd$zenodo_download_PARFs_RRs()

# Download only PARF files (skip RR)
IMPACTncd$zenodo_download_PARFs_RRs(directories = "parf")

# Force re-download (overwrites existing)
IMPACTncd$zenodo_download_PARFs_RRs(overwrite = TRUE)
```

This is useful when:

- A colleague has already generated PARFs/RR tables and you want to reuse
  them instead of re-running the generation step (which can take hours)
- You're setting up a new machine and want to skip time-consuming
  computation

### Downloading Everything at Once

The `zenodo_download_all()` convenience method downloads both input data
**and** simulation outputs in a single call:

```{r download-all}
# New team member ŌĆö connect and download everything
IMPACTncd$zenodo_connect(
  token   = Sys.getenv("ZENODO_SANDBOX_TOKEN"),
  sandbox = TRUE  # concept_doi defaults to "10.5072/zenodo.442996"
)
IMPACTncd$zenodo_download_all()

# Force re-download everything
IMPACTncd$zenodo_download_all(overwrite = TRUE)
```

This calls `zenodo_download_inputs()` followed by
`zenodo_download_PARFs_RRs()` internally. Input data goes to `./inputs/`
and simulation data goes to `./simulation/`.

## Advanced: Archive Control

### Exclude Patterns

Control which directories and files are included in archives:

```{r exclude-patterns}
IMPACTncd$zenodo_upload_inputs(
  version = "1.0.0",
  title   = "...", description = "...", creators = list(...),

  # Skip entire directories matching these regex patterns
  exclude_patterns = c(
    "^unprocessed$",  # Exact match: skip "unprocessed" directory
    "_backup$",       # Ends with "_backup"
    "_old$",          # Ends with "_old"
    "scripts$",       # Ends with "scripts"
    "validation$"     # Ends with "validation"
  ),

  # Skip individual files matching these regex patterns
  exclude_file_patterns = c(
    "\\.R$",     # R scripts
    "\\.Rmd$",   # R Markdown files
    "\\.yaml$",  # YAML config files
    "\\.html$"   # HTML reports
  )
)
```

The default exclude patterns are designed for the IMPACTncd project:

- **Directory patterns**: `^unprocessed$`, `_backup$`, `_old$`, `scripts$`,
  `validation$`
- **Input file patterns**: `\\.R$`, `\\.Rmd$`
- **Simulation file patterns**: `\\.R$`, `\\.csv$`

### Group-By-Prefix

When `group_by_prefix = TRUE` (the default), subdirectories sharing a common
prefix are combined into a single archive. This reduces the number of
individual uploads:

```
inputs/disease_burden/
  Ōö£ŌöĆŌöĆ cancer_2020/    ŌöĆŌöÉ
  Ōö£ŌöĆŌöĆ cancer_2021/    ŌöĆŌöżŌöĆŌöĆ disease_burden_cancer.zip
  Ōö£ŌöĆŌöĆ cancer_2022/    ŌöĆŌöś
  Ōö£ŌöĆŌöĆ chd_2020/       ŌöĆŌöÉ
  Ōö£ŌöĆŌöĆ chd_2021/       ŌöĆŌöżŌöĆŌöĆ disease_burden_chd.zip
  Ōö£ŌöĆŌöĆ chd_2022/       ŌöĆŌöś
  ŌööŌöĆŌöĆ stroke_2020/    ŌöĆŌöĆŌöĆ disease_burden_stroke.zip
```

Set `group_by_prefix = FALSE` to create one archive per subdirectory instead.

### Compression Level

The `compression_level` parameter controls the zip compression level (0-9):

| Level | Speed | Size | Use case |
|-------|-------|------|----------|
| 0 | Fastest | Largest | Already-compressed files (qs, fst) |
| 6 | Balanced | Medium | **Default** ŌĆö good for most data |
| 9 | Slowest | Smallest | When bandwidth is limited |

```{r compression}
# Fast archiving for pre-compressed files
IMPACTncd$zenodo_upload_inputs(
  ...,
  compression_level = 0  # store only, no compression
)
```

### Parallel Archive Creation

Archives are created in parallel by default using PSOCK clusters (which are
safe on all platforms, unlike forking). The number of cores defaults to
`clusternumber` from your simulation design YAML:

```{r parallel}
# Use specific number of cores
IMPACTncd$zenodo_upload_inputs(
  ...,
  multicore = TRUE,
  n_cores   = 10L
)

# Disable parallelism (single-threaded)
IMPACTncd$zenodo_upload_inputs(
  ...,
  multicore = FALSE
)
```

### Preview Archives Before Uploading

Create archives without uploading to inspect sizes and contents:

```{r preview-archives}
# Create archives locally (no Zenodo connection needed)
archives <- IMPACTncd$zenodo_create_archives()

# Inspect results
print(archives[, .(
  archive    = basename(archive_path),
  size_mb    = round(size_bytes / 1024^2, 1),
  file_count = file_count
)])
```

### Automatic Git Tracking Management

When `update_gitignore = TRUE` (the default for upload methods), two things
happen automatically:

1. **`.gitignore` update** ŌĆö Archived data files are added to `.gitignore`
   using efficient directory/extension patterns (e.g., `inputs/mortality/*.fst`).
   This prevents new data files from being accidentally committed in the future.

2. **Git index cleanup** ŌĆö Any data files that were *already tracked* by git
   are removed from the git index via `git rm --cached`. This is critical
   because `.gitignore` only prevents *new* untracked files from being staged;
   files that were previously committed continue to be tracked regardless of
   `.gitignore` rules. The `--cached` flag means files are **kept on disk** ŌĆö
   only the git tracking is removed.

After uploading to Zenodo, your data files live in two places: on your local
disk (for the model to use) and on Zenodo (for reproducibility and sharing).
Anyone cloning the repository will need to run `zenodo_download_inputs()` or
`zenodo_download_PARFs_RRs()` to get the data files.

## Input Directory Structure

The pipeline archives each top-level directory in `./inputs/` separately:

```
inputs/
Ōö£ŌöĆŌöĆ exposure_distributions/    -> exposure_distributions.zip
Ōöé   Ōö£ŌöĆŌöĆ bmi/
Ōöé   Ōö£ŌöĆŌöĆ smoking/
Ōöé   ŌööŌöĆŌöĆ ...
Ōö£ŌöĆŌöĆ disease_burden/            -> disease_burden_*.zip (grouped by prefix)
Ōöé   Ōö£ŌöĆŌöĆ chd_incd/
Ōöé   Ōö£ŌöĆŌöĆ stroke_prvl/
Ōöé   ŌööŌöĆŌöĆ ...
Ōö£ŌöĆŌöĆ mortality/                 -> mortality.zip
Ōö£ŌöĆŌöĆ RR/                        -> RR.zip
Ōö£ŌöĆŌöĆ pop_estimates_lsoa/        -> pop_estimates_lsoa.zip
Ōö£ŌöĆŌöĆ pop_projections/           -> pop_projections.zip
ŌööŌöĆŌöĆ aux_data/                  -> aux_data.zip

simulation/
Ōö£ŌöĆŌöĆ parf/                      -> parf_*.zip (grouped by disease prefix)
Ōöé   Ōö£ŌöĆŌöĆ PARF_af_<hash>.qs
Ōöé   Ōö£ŌöĆŌöĆ PARF_chd_<hash>.qs
Ōöé   ŌööŌöĆŌöĆ ...
ŌööŌöĆŌöĆ rr/                        -> rr.zip
    Ōö£ŌöĆŌöĆ chd.fst
    Ōö£ŌöĆŌöĆ stroke.fst
    ŌööŌöĆŌöĆ ...
```

Each archive contains the full directory structure, so extraction recreates
the correct paths.

## Testing with Sandbox

For testing, use the Zenodo sandbox environment. When you call
`zenodo_connect(sandbox = TRUE)`, the project's sandbox concept DOI
(`10.5072/zenodo.442996`) is set automatically:

```{r sandbox}
# Sandbox ŌĆö concept DOI defaults to 10.5072/zenodo.442996
IMPACTncd$zenodo_connect(
  token   = Sys.getenv("ZENODO_SANDBOX_TOKEN"),
  sandbox = TRUE
)

# Override the default sandbox DOI if needed
IMPACTncd$zenodo_connect(
  token       = Sys.getenv("ZENODO_SANDBOX_TOKEN"),
  sandbox     = TRUE,
  concept_doi = "10.5072/zenodo.999999"
)
```

Note: The sandbox requires a separate account and token from
[sandbox.zenodo.org](https://sandbox.zenodo.org/).

## The ZenodoAssetManager Class

Under the hood, the Simulation class delegates to a `ZenodoAssetManager`
R6 class. You can use it directly for more control:

```{r direct-manager}
library(IMPACTncdEngland)

# Create manager directly
zam <- ZenodoAssetManager$new(
  hash_file = "./simulation/zenodo_manifest.csv",
  logs      = TRUE,
  sandbox   = TRUE
)

# Connect
zam$connect(token = Sys.getenv("ZENODO_SANDBOX_TOKEN"))
zam$set_progress_callback(upload = TRUE, download = TRUE)

# Set concept DOI
zam$set_concept_doi("10.5072/zenodo.442996")

# Get record info
zam$get_record()
print(zam$record)

# List remote files
files <- zam$list_remote_files()
print(files)

# Create archives with full control
archives <- zam$create_input_archives(
  input_base            = "./inputs",
  directories           = c("mortality", "exposure_distributions"),
  exclude_patterns      = c("^unprocessed$"),
  exclude_file_patterns = c("\\.R$", "\\.Rmd$"),
  group_by_prefix       = TRUE,
  compression_level     = 6L,
  multicore             = TRUE,
  n_cores               = 10L,
  update_gitignore      = TRUE
)

# Upload individual archives
for (i in seq_len(nrow(archives))) {
  zam$upload_archive(archives$archive_path[i])
}
```

## Hash Manifests

The pipeline uses file hashes (xxhash64) to detect changes:

```{r manifests}
# Create a manifest of all input files
manifest <- IMPACTncd$zenodo_create_manifest()
print(manifest)

# The ZenodoAssetManager also provides lower-level manifest operations:
zam <- ZenodoAssetManager$new()

manifest <- zam$compute_manifest("./inputs", recursive = TRUE)
zam$save_manifest(manifest)

old_manifest <- zam$load_manifest()
diff <- zam$compare_manifests(manifest, old_manifest)
print(diff$modified)  # Files that changed
print(diff$added)     # New files
print(diff$removed)   # Deleted files
```

## Troubleshooting

### Common Issues

**1. "Not connected to Zenodo"**

```r
# Solution: Call zenodo_connect() first
IMPACTncd$zenodo_connect(sandbox = TRUE)
```

**2. "Zenodo token not provided"**

```r
# Solution: Set environment variable or pass token directly
Sys.setenv(ZENODO_SANDBOX_TOKEN = "your_token")
IMPACTncd$zenodo_connect(
  token   = Sys.getenv("ZENODO_SANDBOX_TOKEN"),
  sandbox = TRUE
)
```

**3. "Record not found for DOI"**

```r
# Solution: Verify DOI format (should be the concept DOI, not a URL)
# Correct: "10.5072/zenodo.442996"
# Wrong:   "https://doi.org/10.5072/zenodo.442996"
IMPACTncd$zenodo_set_doi("10.5072/zenodo.442996")
```

**4. HTTP 403 Forbidden on file listing**

This can happen when zen4R's internal URL does not match the Zenodo server.
The `list_remote_files()` method automatically falls back to a direct API
query. If you still get 403, ensure your token has the correct scopes
(`deposit:write`, `deposit:actions`).

**5. Empty file listing after upload**

zen4R's `listFiles(pretty = TRUE)` can return empty results for **draft**
records on the new Zenodo InvenioRDM backend. The `zenodo_list_files()` method
handles this automatically by falling back to a direct API query to
`/api/records/{id}/draft/files`.

**6. "Missing publisher field required for DOI registration"**

This error occurs when publishing. The `publisher` metadata field is required
by Zenodo's InvenioRDM backend for DOI registration. It defaults to
`"Zenodo"` in `zenodo_upload_inputs()`, so this should only happen if you
created the record manually via the ZenodoAssetManager.

```r
# Fix: Set publisher on the record before publishing
manager$record$setPublisher("Zenodo")
```

**7. Upload shows progress but fails silently**

zen4R may show "No bucket link" warnings and `x` markers during upload, but
the files may actually be uploaded successfully. Always verify with
`zenodo_list_files()` after uploading.

### Network Issues

For slow connections or large files:

```r
# Download specific directories first (start with smaller ones)
IMPACTncd$zenodo_download_inputs(
  directories = c("mortality", "RR")
)
```

## Complete Example Workflow

### Initial Setup (First-Time Upload)

```{r workflow-new}
library(IMPACTncdEngland)

# 1. Create simulation object
IMPACTncd <- Simulation$new("./inputs/sim_design.yaml")

# 2. Connect to Zenodo sandbox for testing
IMPACTncd$zenodo_connect(
  token       = Sys.getenv("ZENODO_SANDBOX_TOKEN"),
  sandbox     = TRUE,
  concept_doi = NULL  # no existing record
)

# 3. (Optional) Preview archives first
archives <- IMPACTncd$zenodo_create_archives()
print(archives[, .(archive = basename(archive_path),
                    size_mb = round(size_bytes / 1024^2, 1))])

# 4. Upload everything
IMPACTncd$zenodo_upload_all(
  title       = "IMPACTncd England Input Data",
  description = paste(
    "Input and simulation data for the IMPACTncd England",
    "microsimulation model, including exposure distributions,",
    "disease burden estimates, mortality data, PARFs, and",
    "compiled relative risk tables."
  ),
  creators = list(
    list(firstname = "Chris", lastname = "Kypridemos",
         orcid = "0000-0002-0746-9229")
  ),
  version = "1.0.0"
)

# 5. Verify upload
files <- IMPACTncd$zenodo_list_files()
cat("Uploaded", nrow(files), "files\n")

# 6. Publish when satisfied (IRREVERSIBLE!)
IMPACTncd$zenodo_publish()
```

### New Team Member (Downloading Data)

```{r workflow-download}
library(IMPACTncdEngland)
IMPACTncd <- Simulation$new("./inputs/sim_design.yaml")

# Connect with the project's concept DOI
IMPACTncd$zenodo_connect(
  token   = Sys.getenv("ZENODO_SANDBOX_TOKEN"),
  sandbox = TRUE  # concept_doi defaults to "10.5072/zenodo.442996"
)

# Check what's available
status <- IMPACTncd$zenodo_check_inputs()

# Download everything (inputs + pre-computed PARFs + RR tables)
IMPACTncd$zenodo_download_all()

# Or download only inputs (if you want to generate PARFs/RR yourself)
# IMPACTncd$zenodo_download_inputs()

# Run simulation
IMPACTncd$run(mc = 1:10, multicore = TRUE, scenario_nam = "sc0")
```

### Publishing Updated Data (New Version)

```{r workflow-update}
# Connect with existing concept DOI
IMPACTncd$zenodo_connect(
  token   = Sys.getenv("ZENODO_SANDBOX_TOKEN"),
  sandbox = TRUE  # concept_doi defaults to "10.5072/zenodo.442996"
)

# Upload as new version (creates new version of existing record)
IMPACTncd$zenodo_upload_all(version = "1.1.0")

# Verify and publish
IMPACTncd$zenodo_list_files()
IMPACTncd$zenodo_publish()

# Check version history
IMPACTncd$zenodo_get_versions()
```

## API Reference

### Simulation Class Methods

| Method | Description |
|--------|-------------|
| `zenodo_connect(token, concept_doi, sandbox, archive_dir, progress)` | Connect to Zenodo and set concept DOI |
| `zenodo_set_doi(concept_doi)` | Change the concept DOI after connecting |
| `zenodo_check_inputs(input_base, directories)` | Check local vs remote sync status |
| `zenodo_download_inputs(input_base, directories, overwrite)` | Download and extract input archives |
| `zenodo_download_PARFs_RRs(simulation_base, directories, overwrite)` | Download and extract simulation archives |
| `zenodo_download_all(input_base, ..., simulation_base, ..., overwrite)` | Download inputs + simulation in one call |
| `zenodo_upload_inputs(input_base, ..., publish)` | Create archives and upload inputs |
| `zenodo_upload_simulation(simulation_base, directories, ...)` | Upload simulation outputs (PARF, RR) |
| `zenodo_upload_all(input_base, simulation_base, ..., publish)` | Upload inputs + simulation in one call |
| `zenodo_publish()` | Publish the draft record (**irreversible!**) |
| `zenodo_list_files(concept_doi)` | List all files in the current record |
| `zenodo_get_versions()` | Get version history with DOIs |
| `zenodo_create_archives(input_base, ...)` | Create archives locally (no upload) |
| `zenodo_create_manifest(input_base, save)` | Create file hash manifest |

### Archive Creation Parameters

These parameters are available in `zenodo_upload_inputs()`,
`zenodo_upload_all()`, and `zenodo_create_archives()`:

| Parameter | Default | Description |
|-----------|---------|-------------|
| `exclude_patterns` | `c("^unprocessed$", "_backup$", ...)` | Regex patterns for directories to skip |
| `exclude_file_patterns` | `c("\\.R$", "\\.Rmd$")` | Regex patterns for files to skip |
| `group_by_prefix` | `TRUE` | Combine subdirectories by shared prefix |
| `compression_level` | `6` | Zip compression (0=store, 9=max) |
| `multicore` | `TRUE` | Parallel archive creation |
| `n_cores` | `clusternumber` from YAML | Number of CPU cores |
| `update_gitignore` | `TRUE` | Auto-update .gitignore and remove tracked data files from git index |

## References

- [Zenodo](https://zenodo.org/) - Open repository hosted by CERN
- [zen4R package](https://github.com/eblondel/zen4R) - R interface to Zenodo
  API
- [Zenodo REST API](https://developers.zenodo.org/) - API documentation
- [DOI Foundation](https://www.doi.org/) - About Digital Object Identifiers
