---
title: "Managing Input Data with Zenodo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Managing Input Data with Zenodo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

IMPACTncdEngland includes a robust pipeline for managing input data files via [Zenodo](https://zenodo.org/), a general-purpose open repository. This enables:

- **Reproducibility**: DOI-based versioning ensures you always know which data version was used
- **Collaboration**: Easy sharing of input datasets across team members
- **Archiving**: Long-term preservation of input data with persistent identifiers
- **Version Control**: Track changes to input data over time

The pipeline uses the `zen4R` package to interact with the Zenodo REST API, and supports both **parquet** and **fst** file formats commonly used in the simulation.

## Prerequisites

### 1. Install Required Packages

The following packages are required:

```{r install-packages}
install.packages(c("zen4R", "zip", "digest"))
```

### 2. Get a Zenodo Access Token

1. Create a Zenodo account at [zenodo.org](https://zenodo.org/)
2. Go to [Applications > Personal access tokens](https://zenodo.org/account/settings/applications/tokens/new/)
3. Create a new token with **deposit:write** and **deposit:actions** scopes
4. Save the token securely (see below)

### 3. Secure Token Management

Your Zenodo token grants full access to your Zenodo deposits. **Never commit tokens to version control!**

#### Option A: `.Renviron` File (Recommended for Personal Use)

The `.Renviron` file stores environment variables that R loads at startup:

```r
# Open .Renviron for editing
usethis::edit_r_environ()
```

Add this line to the file:
```
ZENODO_TOKEN=your_token_here
```

Save and restart R. The token is now available via `Sys.getenv("ZENODO_TOKEN")`.

**Security considerations:**

- The `.Renviron` file is stored in your home directory (`~/.Renviron`)
- It is **not** in your project folder, so it won't be accidentally committed to git
- Set restrictive permissions: `chmod 600 ~/.Renviron` (Linux/macOS)

#### Option B: `keyring` Package (Most Secure)
 
The `keyring` package stores secrets in your operating system's secure credential store
(Windows Credential Manager, macOS Keychain, or Linux Secret Service):

```r
# Install keyring
install.packages("keyring")

# Store the token (you'll be prompted to enter it securely)
keyring::key_set("zenodo", "token")

# Retrieve the token in your scripts
token <- keyring::key_get("zenodo", "token")
IMPACTncd$zenodo_connect(token = token)
```

**Advantages:**

- Tokens are encrypted by the OS
- Works across R sessions without configuration files
- Integrates with system authentication (fingerprint, password, etc.)

#### Option C: Environment Variable in Session (Temporary)

For one-time use, set the token in your current R session only:

```r
Sys.setenv(ZENODO_TOKEN = "your_token_here")
```

**Note:** This does not persist after the session ends and requires re-entry each time.

#### Option D: CI/CD Secrets (For Automated Workflows)

For GitHub Actions or other CI/CD systems:

1. Add `ZENODO_TOKEN` as a repository secret in GitHub Settings > Secrets
2. Reference it in your workflow:

```yaml
env:
  ZENODO_TOKEN: ${{ secrets.ZENODO_TOKEN }}
```

#### Security Best Practices

| Do | Don't |
|-----|--------|
| ✅ Use `.Renviron` or `keyring` | ❌ Hardcode tokens in scripts |
| ✅ Add `.Renviron` to `.gitignore` if project-local | ❌ Commit tokens to git |
| ✅ Use separate tokens for sandbox vs production | ❌ Share tokens between users |
| ✅ Regenerate tokens periodically | ❌ Use tokens with excessive permissions |
| ✅ Revoke tokens you no longer need | ❌ Log or print tokens in output |

#### Verifying Your Setup

```r
# Check if token is available (without printing it!)
if (nzchar(Sys.getenv("ZENODO_TOKEN"))) {
  message("Zenodo token is configured")
} else {
  message("Zenodo token is NOT configured")
}
```

## Quick Start

### Downloading Input Data

If you're setting up the model for the first time, download the input data from Zenodo:

```{r download-example}
library(IMPACTncdEngland)

# Create simulation object
IMPACTncd <- Simulation$new("./inputs/sim_design.yaml")

# Connect to Zenodo and set the concept DOI
IMPACTncd$zenodo_connect(concept_doi = "10.5281/zenodo.XXXXXXX")

# Check what's available
IMPACTncd$zenodo_check_inputs()

# Download all missing inputs
IMPACTncd$zenodo_download_inputs()

# Or download specific directories
IMPACTncd$zenodo_download_inputs(directories = c("exposure_distributions", "disease_burden"))
```

### Uploading Input Data

To share your input data via Zenodo:
        
```{r upload-example}
# Connect to Zenodo
IMPACTncd$zenodo_connect()

# For a NEW record (first time):
IMPACTncd$zenodo_upload_inputs(
  title = "IMPACTncd England Input Data",
  description = "Input data files for the IMPACTncd England microsimulation model, including exposure distributions, disease burden estimates, and mortality data.",
  creators = list(
    list(
      firstname = "Chris",
      lastname = "Kypridemos", 
      orcid = "0000-0002-0746-9229"
    )
  ),
  version = "1.0.0"
)

# For a NEW VERSION of existing record:
IMPACTncd$zenodo_connect(concept_doi = "10.5281/zenodo.XXXXXXX")
IMPACTncd$zenodo_upload_inputs(version = "1.1.0")

# When ready, publish the record (IRREVERSIBLE!)
IMPACTncd$zenodo_publish()
```

## Detailed Usage

### The ZenodoAssetManager Class

Under the hood, the Simulation class uses a `ZenodoAssetManager` R6 class that handles all Zenodo operations. You can also use it directly for more control:

```{r direct-manager}
library(IMPACTncdEngland)

# Create manager directly
zam <- ZenodoAssetManager$new(
  hash_file = "./simulation/zenodo_manifest.csv",
  logs = TRUE,
  sandbox = FALSE  # Set TRUE for testing
)

# Connect
zam$connect()

# Set concept DOI
zam$set_concept_doi("10.5281/zenodo.XXXXXXX")

# Get record info
record <- zam$get_record()
print(record)

# List remote files
files <- zam$list_remote_files()
print(files)

# Get all versions
versions <- zam$get_versions()
print(versions)
```

### Creating Archives

The pipeline groups files into zip archives by directory for efficient storage.
Zip format is used for cross-platform compatibility (works on Linux, Windows, and macOS):

```{r create-archives}
# Create zip archives for all input directories
archives <- zam$create_input_archives(input_base = "./inputs")
print(archives)

# Create archive for specific directory
archive_path <- zam$create_archive(
  input_dir = "./inputs/exposure_distributions",
  archive_name = "exposure_distributions"
)
```

### Hash Manifests

The pipeline uses file hashes to detect changes:

```{r manifests}
# Create a manifest of all input files
manifest <- zam$compute_manifest(
  input_dir = "./inputs",
  recursive = TRUE
)
print(manifest)

# Save manifest
zam$save_manifest(manifest)

# Load existing manifest
old_manifest <- zam$load_manifest()

# Compare manifests
diff <- zam$compare_manifests(manifest, old_manifest)
print(diff$modified)  # Files that changed
print(diff$added)     # New files
print(diff$removed)   # Deleted files
```

## Input Directory Structure

The pipeline archives each top-level directory in `./inputs/` separately:

```
inputs/
├── exposure_distributions/    → exposure_distributions.zip
│   ├── bmi/
│   ├── smoking/
│   └── ...
├── disease_burden/            → disease_burden.zip
│   ├── chd_incd/
│   ├── stroke_prvl/
│   └── ...
├── mortality/                 → mortality.zip
├── RR/                        → RR.zip
├── pop_estimates_lsoa/        → pop_estimates_lsoa.zip
├── pop_projections/           → pop_projections.zip
└── aux_data/                  → aux_data.zip
```

Each archive contains the full directory structure, so extraction recreates the correct paths.

## Version Control with DOIs

Zenodo provides two types of DOIs:

1. **Concept DOI**: Resolves to the *latest* version of a record. Use this for sharing.
2. **Version DOI**: Points to a *specific* version. Use this for reproducibility.

```{r dois}
# Get both DOIs
record <- zam$get_record()
concept_doi <- record$conceptdoi      # e.g., 10.5281/zenodo.1234567
version_doi <- record$pids$doi$identifier  # e.g., 10.5281/zenodo.1234568

# Get version history
versions <- zam$get_versions()
# Returns: date, version, doi for each version
```

### Best Practices for Versioning

1. **Use semantic versioning**: `MAJOR.MINOR.PATCH`
   - MAJOR: Breaking changes (e.g., restructured directories)
   - MINOR: New data added (e.g., new disease)
   - PATCH: Bug fixes (e.g., corrected values)

2. **Document changes**: Update the record description with changelog

3. **Pin versions in publications**: Always cite the version DOI, not concept DOI

## Testing with Sandbox

For testing, use the Zenodo sandbox environment:

```{r sandbox}
# Connect to sandbox
IMPACTncd$zenodo_connect(sandbox = TRUE)

# Or with manager directly
zam <- ZenodoAssetManager$new(sandbox = TRUE)
zam$connect(token = Sys.getenv("ZENODO_SANDBOX_TOKEN"))
```

Note: The sandbox requires a separate account and token from [sandbox.zenodo.org](https://sandbox.zenodo.org/).

## Handling Conflicts

When local files differ from remote:

```{r conflicts}
# Check status first
status <- IMPACTncd$zenodo_check_inputs()
print(status)

# If local directories exist but differ, you can:

# Option 1: Keep local (do nothing)
IMPACTncd$zenodo_download_inputs(overwrite = FALSE)  # Skips existing

# Option 2: Overwrite with remote
IMPACTncd$zenodo_download_inputs(overwrite = TRUE)

# Option 3: Backup local and download
# (Manual process)
file.rename("./inputs/disease_burden", "./inputs/disease_burden_backup")
IMPACTncd$zenodo_download_inputs(directories = "disease_burden")
```

## Cross-Platform Compatibility

The pipeline uses the `zip` format exclusively for cross-platform compatibility:

- **Linux**: Uses the `zip` R package
- **macOS**: Uses the `zip` R package  
- **Windows**: Uses the `zip` R package
- **Docker**: Works in containerized environments

The `zip` package is a pure R implementation that works identically on all platforms,
ensuring that archives created on one system can be extracted on another.

## Troubleshooting

### Common Issues

**1. "Not connected to Zenodo"**
```{r}
# Solution: Call connect() first
IMPACTncd$zenodo_connect()
```

**2. "Zenodo token not provided"**
```{r}
# Solution: Set environment variable or pass token directly
Sys.setenv(ZENODO_TOKEN = "your_token")
IMPACTncd$zenodo_connect()
# Or
IMPACTncd$zenodo_connect(token = "your_token")
```

**3. "Record not found for DOI"**
```{r}
# Solution: Verify DOI format (should be the concept DOI)
# Correct: "10.5281/zenodo.1234567"
# Wrong:   "https://doi.org/10.5281/zenodo.1234567"
IMPACTncd$zenodo_set_doi("10.5281/zenodo.1234567")
```

**4. Upload fails with large files**
```{r}
# Zenodo has a 50GB limit per record
# Consider splitting into multiple records for very large datasets
```

### Network Issues

For slow connections or large files:

```{r}
# The zen4R package handles retries automatically
# For persistent issues, try downloading specific directories
IMPACTncd$zenodo_download_inputs(
  directories = c("mortality", "RR")  # Start with smaller directories
)
```

## Complete Example Workflow

### Initial Setup (New Project)

```{r workflow-new}
library(IMPACTncdEngland)

# 1. Create simulation object
IMPACTncd <- Simulation$new("./inputs/sim_design.yaml")

# 2. Connect to Zenodo
IMPACTncd$zenodo_connect(concept_doi = "10.5281/zenodo.XXXXXXX")

# 3. Check available data
status <- IMPACTncd$zenodo_check_inputs()

# 4. Download all inputs
IMPACTncd$zenodo_download_inputs()

# 5. Run simulation
IMPACTncd$run(mc = 1:10, multicore = TRUE, scenario_nam = "sc0")
```

### Publishing Updated Data

```{r workflow-update}
# 1. Connect with existing DOI
IMPACTncd$zenodo_connect(concept_doi = "10.5281/zenodo.XXXXXXX")

# 2. Create manifest to track changes
manifest <- IMPACTncd$zenodo_create_manifest()

# 3. Upload new version
IMPACTncd$zenodo_upload_inputs(version = "1.2.0")

# 4. Review at Zenodo web interface
# 5. When ready, publish
IMPACTncd$zenodo_publish()
```

## API Reference

### Simulation Class Methods

| Method | Description |
|--------|-------------|
| `zenodo_connect(token, concept_doi, sandbox)` | Connect to Zenodo API |
| `zenodo_set_doi(concept_doi)` | Set the concept DOI for the record |
| `zenodo_check_inputs(input_base, directories)` | Check sync status |
| `zenodo_download_inputs(input_base, directories, overwrite)` | Download data |
| `zenodo_upload_inputs(input_base, directories, version, ...)` | Upload data |
| `zenodo_publish()` | Publish the record (irreversible!) |
| `zenodo_get_versions()` | Get version history |
| `zenodo_create_manifest(input_base, save)` | Create file hash manifest |

### ZenodoAssetManager Class

See `?ZenodoAssetManager` for full documentation.

## References

- [Zenodo](https://zenodo.org/) - Open repository
- [zen4R package](https://github.com/eblondel/zen4R) - R interface to Zenodo API
- [Zenodo REST API](https://developers.zenodo.org/) - API documentation
